---
title: "Indicator Species Analysis Boxplot"
author: Yanting Xiong
output: html_document
date: "2026-03-20"
---
When plotting, a stat cutoff of greater than and equal to 0.35 is used, therefore indicator taxa with a stat value of below 0.35 will not be plotted.

## Library loading

```{r library_loading}
library(tidyverse)
library(phyloseq)
library(indicspecies)
library(dplyr)
library(ggpubr)
```

## Dataset loading
```{r dataset_loading}
# Load in the filtered and wrangled Parkinson's metadata
parkinsons_metadata_filt_cat = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# Relevel the categorical variable in the metadata so that it appears in the order “low,” “medium,” and “high” in the plot, rather than the default alphabetical order
vars = c("Tert_BC", "Tert_AC", "Tert_VA", "Tert_VC", "Tert_VE", "Tert_combine") # store the column names

prefix = c("BC", "AC", "VA", "VC", "VE", "Combine") # stores the label prefixes used in the factor levels

for (i in seq_along(vars)) {
  parkinsons_metadata_filt_cat[[vars[i]]] <- fct_relevel(
    parkinsons_metadata_filt_cat[[vars[i]]],
    paste(prefix[i], "Low"),
    paste(prefix[i], "Medium"),
    paste(prefix[i], "High")
  )
}

# Load in feature table exported from QIIME2
counts = read.delim('../../../QIIME2_outputs/feature-table.txt', skip = 1, row.names = 1)

# Load in taxonomy table exported from QIIME2
taxonomy = read.delim('../../../QIIME2_outputs/taxonomy.tsv', row.names = 1)
```

## Create a Phyloseq Object

```{r phyloseq_object}
# Taxonomy
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon, 
           into = c("Domain","Phylum","Class","Order",
                    "Family","Genus","Species"),
           sep = ";", fill = "right") %>% 
  select(-Confidence) %>% 
  as.matrix()

# Counts
counts_formatted = counts %>% as.matrix()

# Create the phyloseq object 
ps = phyloseq(sample_data(parkinsons_metadata_filt_cat),
              otu_table(counts_formatted, taxa_are_rows = T),
              tax_table(taxonomy_formatted))

```

## Preprocessing

```{r preprocessing}
# Aggregate ASVs to higher taxonomic level (genus level in this case)
ps_genus = tax_glom(ps, "Genus")

# Convert phyloseq to relative abundance
ps_relab = microbiome::transform(ps_genus, "compositional")

# Extract the otu_table as a data frame because only the otu_table, instead of the phyloseq object, will be used to make plots
otu_table = data.frame(otu_table(ps_relab))

# Extract all the feature IDs from the otu_table
feature_id_extract = rownames(otu_table)

```

## Beta Carotene

### Indicator taxa in the high intake group

#### Cloacibacillus 

``` {r boxplot_for_Cloacibacillus}
phyla_to_plot_BC_high_Cloa = feature_id_extract[feature_id_extract %in% "630532cabb74720c1204579e6fe741b4"] 

df_of_taxa_BC_high_Cloa = prune_taxa(phyla_to_plot_BC_high_Cloa, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001313370

df_of_taxa_BC_high_Cloa %>% 
  ggplot(aes(x = Tert_BC, y = log10(Abundance + 0.0001313370/2), fill = Tert_BC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Beta-carotene Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

#### Alloprevotella

``` {r boxplot_for_Alloprevotella}
# Extract the tax table
tax_table = as.data.frame(tax_table(ps))

# Find the Genus for the ASV of interest
tax_table["a030e5d5ff9c75a027e6cc58ff02cf7c", "Genus"] # g_Alloprevotella

# Extract the table containing aggregated ASVs
tax_table_genus = as.data.frame(tax_table(ps_genus))

# Set the target genus
target_genus = "g__Alloprevotella"

# Clean the column
tax_table_genus$Genus <- as.character(tax_table_genus$Genus)

tax_table_genus$Genus <- trimws(tax_table_genus$Genus)

# Find which ASV name is being used as the representative
representative_asv = rownames(tax_table_genus)[tax_table_genus$Genus == "g__Alloprevotella"]
representative_asv # cb645d9e303b0e2d761969abb610d210

phyla_to_plot_BC_high_Allo = feature_id_extract[feature_id_extract %in% "cb645d9e303b0e2d761969abb610d210"] 

df_of_taxa_BC_high_Allo = prune_taxa(phyla_to_plot_BC_high_Allo, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001310101

df_of_taxa_BC_high_Allo %>% 
  ggplot(aes(x = Tert_BC, y = log10(Abundance + 0.0001310101/2), fill = Tert_BC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Beta-carotene Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

### Indicator taxa in the low intake group

#### Escherichia-Shigella

``` {r boxplot_for_Escherichia-Shigella_first}
phyla_to_plot_BC_low_Esch = feature_id_extract[feature_id_extract %in% "4c8288bfbd76958c0c094d87b97650f8"] 

df_of_taxa_BC_low_Esch = prune_taxa(phyla_to_plot_BC_low_Esch, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001518834

# Create p-value data frame
stat.test_BC = data.frame(
  group1 = "BC Low",
  group2 = "BC High",
  p = "0.006",
  y.position = 0
)

BC_low_Esch_plot = df_of_taxa_BC_low_Esch %>% 
  ggplot(aes(x = Tert_BC, y = log10(Abundance + 0.0001518834/2), fill = Tert_BC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_manual(
    name = "Beta-carotene Intake Categorization",
    values = c("BC Low" = "steelblue",
               "BC Medium" = "goldenrod2",
               "BC High" = "indianred3"),
    labels = c("Low intake", "Medium intake", "High intake")
  ) +
  stat_pvalue_manual(stat.test_BC, label = "p")
BC_low_Esch_plot

# Export the plot
ggsave("../../../Results/Figures/Aim_2/Indicator_species/Relative Abundance of Escherichia–Shigella Across Beta-Carotene Intake Levels in PD.png", plot = BC_low_Esch_plot, width = 7, height = 8, dpi = 300)
```

## Vitamin A

### Indicator taxa in the high intake group

#### Acidaminococcus

``` {r boxplot_for_Acidaminococcus}
phyla_to_plot_VA_high_Acida = feature_id_extract[feature_id_extract %in% "7bf7ea59cac793e65c0bc293eabf44dc"] 

df_of_taxa_VA_high_Acida = prune_taxa(phyla_to_plot_VA_high_Acida, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0006432050

df_of_taxa_VA_high_Acida %>% 
  ggplot(aes(x = Tert_VA, y = log10(Abundance + 0.0006432050/2), fill = Tert_VA)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin A Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

## Vitamin C

### Indicator taxa in the high intake group

#### Anaerosporobacter

``` {r boxplot_for_Anaerosporobacter}
phyla_to_plot_VC_high_Anae = feature_id_extract[feature_id_extract %in% "a3d1f4355f204d4d49ad2bea3b801265"] 

df_of_taxa_VC_high_Anae = prune_taxa(phyla_to_plot_VC_high_Anae, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001980786

df_of_taxa_VC_high_Anae %>% 
  ggplot(aes(x = Tert_VC, y = log10(Abundance + 0.0001980786/2), fill = Tert_VC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin C Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

#### Asteroleplasma

``` {r boxplot_for_Asteroleplasma}
phyla_to_plot_VC_high_Astero = feature_id_extract[feature_id_extract %in% "6ca3f1e7d3011097ec41585e61307d26"] 

df_of_taxa_VC_high_Astero = prune_taxa(phyla_to_plot_VC_high_Astero, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0002391582

df_of_taxa_VC_high_Astero %>% 
  ggplot(aes(x = Tert_VC, y = log10(Abundance + 0.0002391582/2), fill = Tert_VC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin C Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

### Indicator taxa in the low intake group

#### Escherichia-Shigella

``` {r boxplot_for_Escherichia-Shigella_second}
phyla_to_plot_VC_low_Esch = feature_id_extract[feature_id_extract %in% "4c8288bfbd76958c0c094d87b97650f8"] 

df_of_taxa_VC_low_Esch = prune_taxa(phyla_to_plot_VC_low_Esch, ps_relab) %>% psmelt()

# Create p-value data frame
stat.test_VC = data.frame(
  group1 = "VC Low",
  group2 = "VC High",
  p = "0.003",
  y.position = 0
)

# The smallest nonzero relative abundance value is 0.0001518834

VC_low_Esch_plot = df_of_taxa_VC_low_Esch %>% 
  ggplot(aes(x = Tert_VC, y = log10(Abundance + 0.0001518834/2), fill = Tert_VC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_manual(
    name = "Vitamin C Intake Categorization",
    values = c("VC Low" = "steelblue",
               "VC Medium" ="goldenrod2",
               "VC High" = "indianred3"),
    labels = c("Low intake", "Medium intake", "High intake")
  ) +
  stat_pvalue_manual(stat.test_VC, label = "p")
VC_low_Esch_plot

# Export the plot
ggsave("../../../Results/Figures/Aim_2/Indicator_species/Relative Abundance of Escherichia–Shigella Across Vitamin C Intake Levels in PD.png", plot = VC_low_Esch_plot, width = 7, height = 8, dpi = 300)
```

#### Acidaminococcus

``` {r boxplot_for_Acidaminococcus}
phyla_to_plot_VC_low_Acida = feature_id_extract[feature_id_extract %in% "7bf7ea59cac793e65c0bc293eabf44dc"] 

df_of_taxa_VC_low_Acida = prune_taxa(phyla_to_plot_VC_low_Acida, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0006432050

df_of_taxa_VC_low_Acida %>% 
  ggplot(aes(x = Tert_VC, y = log10(Abundance + 0.0006432050/2), fill = Tert_VC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin C Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

#### Faecalitalea

``` {r boxplot_for_Faecalitalea}
phyla_to_plot_VC_low_Fae = feature_id_extract[feature_id_extract %in% "f0135ea5c9854954ff851185c476e91c"] 

df_of_taxa_VC_low_Fae = prune_taxa(phyla_to_plot_VC_low_Fae, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001390530

df_of_taxa_VC_low_Fae %>% 
  ggplot(aes(x = Tert_VC, y = log10(Abundance + 0.0001390530/2), fill = Tert_VC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin C Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

## Vitamin E

### Indicator taxa in the high intake group

#### Victivallaceae

``` {r boxplot_for_Victivallaceae}
phyla_to_plot_VE_high_Vic = feature_id_extract[feature_id_extract %in% "b7fb7b8de1b4e013e7255cec428893a7"] 

df_of_taxa_VE_high_Vic = prune_taxa(phyla_to_plot_VE_high_Vic, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0001997470

df_of_taxa_VE_high_Vic %>% 
  ggplot(aes(x = Tert_VE, y = log10(Abundance + 0.0001997470/2), fill = Tert_VE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin E Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

### Indicator taxa in the low intake group

#### [Ruminococcus] gnavus group

``` {r boxplot_for_[Ruminococcus]_gnavus_group}
phyla_to_plot_VE_low_Rum = feature_id_extract[feature_id_extract %in% "93f6e99ff381954b541d29cc24c2be7c"] 

df_of_taxa_VE_low_Rum = prune_taxa(phyla_to_plot_VE_low_Rum, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0004671650

df_of_taxa_VE_low_Rum %>% 
  ggplot(aes(x = Tert_VE, y = log10(Abundance + 0.0004671650/2), fill = Tert_VE)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Vitamin E Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

## Combined

### Indicator taxa in the high intake group

#### Lachnospiraceae FCS020 group

``` {r boxplot_for_Lachnospiraceae_FCS020_group}
phyla_to_plot_combined_high_Lach = feature_id_extract[feature_id_extract %in% "4661cd41d2df6a4bf70276894e6cc779"] 

df_of_taxa_combined_high_Lach = prune_taxa(phyla_to_plot_combined_high_Lach, ps_relab) %>% psmelt()

# The smallest nonzero relative abundance value is 0.0002972946

df_of_taxa_combined_high_Lach %>% 
  ggplot(aes(x = Tert_combine, y = log10(Abundance + 0.0002972946/2), fill = Tert_combine)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2) +
  labs(x = "Dietary Antioxidant Intake Tertile", y = "Log-10 Transformed Relative Abundance") +
  scale_x_discrete(labels = c("Low intake", "Medium intake", "High intake")) +
  scale_fill_discrete(
    name = "Combined Dietary Antioxidant Intake Categorization",
    labels = c("Low intake", "Medium intake", "High intake")
  )

```

