---
title: "rerunning beta carotene with DAI standardization"
author: "Cynthia"
date: "2026-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r}
library(tidyverse)
library(dplyr)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
library(ggpicrust2)
library(Maaslin2)
```

different DAI analysis method

```{r}
metadata = read.delim('../Datasets/parkinsons_metadata.txt', row.names = 1)
taxonomy = read.delim('../Datasets/taxonomy.tsv', row.names = 1) # from qiime2
counts = read.delim('../Datasets/feature-table.txt', skip=1, row.names=1)  # from qiime2
tree = read_tree('../Datasets/tree.nwk')

# filter for only PD
metadata_pd = metadata %>%
  filter(Disease == "PD")
sum(is.na(metadata_pd$Beta_carotene)) 
# 13 NA values in the Beta_carotene column that I will be removing them for this analysis

metadata_pd_no_NA = metadata_pd %>%
  filter(!is.na(Beta_carotene))

# DAI calculation -> $$Standardized\ Score = \frac{\text{Individual Intake} - \text{Mean Intake}}{\text{Standard Deviation}}$$

metadata_pd_no_NA$Beta_carotene_DAI = as.vector(scale(
  metadata_pd_no_NA$Beta_carotene))

head(metadata_pd_no_NA[, c("Beta_carotene", "Beta_carotene_DAI")])

summary(metadata_pd_no_NA$Beta_carotene_DAI)

```
Tertile Approach (High vs. Low Contrast)
Since I found no results before I want to try to see a stronger effect in my moderation model. I will split the data into thirds (Tertiles) and compare the top third to the bottom third. This removes the "middle" group where the signal might be noisy.
```{r}
# Calculate tertile cut-offs
tertiles = quantile(metadata_pd_no_NA$Beta_carotene_DAI, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)

# Categorize into Low, Mid, High
metadata_pd_no_NA$DAI_Tertile <- cut(metadata_pd_no_NA$Beta_carotene_DAI, 
                             breaks = tertiles, 
                             labels = c("Low Intake", "Medium Intake", "High Intake"), 
                             include.lowest = TRUE)

# Filter to compare only High vs. Low in your regression


print(table(metadata_pd_no_NA$DAI_Tertile))

```
Taxonomy and phylo again
```{r}
# Taxonomy
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon, 
           into = c('Domain','Phylum','Class','Order',
                    'Family','Genus','Species'),
           sep=';', fill='right') %>% 
  select(-Confidence) %>% 
  as.matrix()

counts_formatted = counts %>% as.matrix()

# PS object
ps = phyloseq(sample_data(metadata_pd_no_NA),
              otu_table(counts_formatted, taxa_are_rows = T),
              tax_table(taxonomy_formatted),
              tree)

saveRDS(ps,'../Datasets/phyloseq_taxonomy.rds')

```
beta-diversity
```{r}
rare_data = counts
rarecurve(t(rare_data), step = 50, col = "blue", label = FALSE, 
          main = "Rarefaction Curves for PD Cohort",
          xlab = "Sequencing Depth", ylab = "Observed Features")
# A sampling depth of 6000 is chosen because the curves for
# majority of the samples reach a plateau (asymptote) by this point. 

# rarefy the data:
psrare = ps %>% rarefy_even_depth(sample.size = 6000, rngseed = 421)

# Calculate the distance between each sample pair
ps_bray = phyloseq::distance(psrare, method = "bray")
View(as.matrix(ps_bray)) # Every pairwise comparison included

# MDS scaling
set.seed(421)
mds = metaMDS(ps_bray)

# Extract data
mds_data = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data)

# Plot it
beta_plot_beta_carotene_intake = mds_data %>%
  ggplot(aes(MDS1,MDS2,color = DAI_Tertile)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size=18) +
  labs(color = "Beta Carotene Intake")
print(beta_plot_beta_carotene_intake)

```
Permanova 
```{r}
metadata_perm = psrare %>% 
  sample_data() %>% 
  data.frame() %>% 
  rownames_to_column("SampleID")
# Running PERMANOVA
#metadata_perm = sample_data(psrare) %>% data.frame() # to make the next lines easier

# Single variable
stats_univar = adonis2(ps_bray ~ DAI_Tertile, data = metadata_perm)

# Two variables - adding Sex as a covariable, just to see if controlling for sex makes a difference
stats_multivar = adonis2(ps_bray ~ DAI_Tertile + Sex, data = metadata_perm, by = 'margin') 

# Combine into a table
stats = bind_rows('Univariate' = stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  'Multivariate' = stats_multivar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') # First column is Model, either Uni or Multi

# Formatting
stats = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

print(stats)

# Groups to compare
to_compare = c('Low', 'High')

# List samples that belong to those groups
samples_to_keep = metadata_perm %>% 
  filter(DAI_Tertile %in% to_compare) %>% 
  pull(SampleID)

#samples_to_keep = metadata_perm %>% filter(DAI_Tertile %in% to_compare) %>% rownames()

print(length(samples_to_keep))
# Subset the distance matrix to only include those samples
ps_bray_sub = as.matrix(ps_bray)[samples_to_keep,samples_to_keep] %>% as.dist()

# Run a PERMANOVA
stats_univar_final = adonis2(ps_bray_sub ~ DAI_Tertile,
                       data = metadata_perm %>% filter(DAI_Tertile %in% to_compare))

print(stats_univar_final)

```
differencial abundance
```{r}
set.seed(421)

# Tax_glom
ps_glom = tax_glom(ps, 'Genus')

# Removing genera with mean relative abundance < 0.0001
ps_relab = transform_sample_counts(ps_glom, function(x) x / sum(x))

# genera to keep
genera_to_keep = taxa_names(ps_relab)[taxa_sums(ps_relab) / nsamples(ps_relab) >= 0.0001]
ps_filtered = prune_taxa(genera_to_keep, ps_glom)

set.seed(421)
out = ancombc2(data = ps_filtered, 
               fix_formula = 'DAI_Tertile', 
               p_adj_method = 'BH',
               prv_cut = 0.1)

statistical_table = out$res

# Filter stats
genera_filtered = statistical_table %>% 
  filter(`diff_DAI_TertileHigh`==T) 

dim(genera_filtered)

```
Picrust
```{r}
meta = readRDS('../Datasets/phyloseq_taxonomy.rds') %>% .@sam_data %>% data.frame() %>% 
  rownames_to_column('sample_name')

pathway_abundance = read.delim('../Datasets/path_abun_unstrat.tsv',row.names = 1)

# Filter metadata to just High and Low
meta_sub = metadata_pd_no_NA %>% 
  filter(DAI_Tertile %in% c("High Intake", "Low Intake")) %>%
  mutate(DAI_Tertile = factor(DAI_Tertile, levels = c("High Intake", "Low Intake")))

# Ensure the abundance matrix matches the filtered metadata samples
path_sub = path_filt[, rownames(meta_sub)]
# Filter
path_filt = pathway_abundance %>% select(all_of(rownames(metadata_pd_no_NA)))


diff_abun = pathway_daa(abundance = path_filt,
                   metadata = meta_sub,
                   group = "DAI_Tertile",
                   daa_method = "Maaslin2",
                   select = NULL,
                   reference = "High Intake")

daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = diff_abun,
                                              ko_to_kegg = FALSE)
 
# save
saveRDS(daa_annotated_results_df,'daa_annotated_results_df.rds')

readRDS('daa_annotated_results_df.rds')
```
Plotting
```{r}
source('ggpicrust2_errorbar_function_fixed.R')

# Filter out NA in p-values and log2 fold change columns
daa_annotated_results_df_filt = daa_annotated_results_df %>%
  filter(!is.na(p_adjust), !is.na(log2_fold_change))

daa_annotated_results_df_filt = daa_annotated_results_df%>%
  filter(!is.na(description))

# Adjusted P-value = 32 hits
daa_annotated_results_df_filt %>% 
  filter(p_adjust < 0.04) %>% 
  nrow()

# Log2 Fold Change values = 8 hits (sufficient for plot)
daa_annotated_results_df_filt %>% 
  filter(abs(log2_fold_change) >= 5e-3) %>% 
  nrow()

daa_annotated_results_df_filt %>% filter(p_adjust< 0.05, abs(log2_fold_change) > 5e-3) %>% nrow()

peb = pathway_errorbar_fixed(abundance = path_filt, 
                     daa_results_df = daa_annotated_results_df_filt, 
                     Group = metadata_pd_no_NA$DAI_Tertile, 
                     wrap_label = T, wraplength=60,
                     fc_cutoff = 5e-3, order_by_log = F,
                     p_values_threshold = 0.04, 
                     order = "name", 
                     ko_to_kegg = FALSE, 
                     p_value_bar = TRUE, 
                     x_lab = "description")

ggsave("peb Error Bar Fixed.png",
       plot = peb, width = 18, height = 10)
```






## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
