---
title: "Diversity Analysis"
output: html_document
---

```{r libraries, include = FALSE}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
```

## Load the data

```{r load_data}
parkinsons_metadata_filt_cat = read.csv('../Datasets/Metadata/parkinsons_metadata_filt_cat.csv', row.names = 1)
counts = read.delim('../QIIME2_outputs/feature-table.txt', skip = 1, row.names = 1)
taxonomy = read.delim('../QIIME2_outputs/taxonomy.tsv', row.names = 1)
```

## Make a Phyloseq object

```{r Create a Phyloseq object}
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon,
           into = c("Domain", "Phylum", "Class", "Order", "Family",
                    "Genus", "Species"),
           sep = ";", fill = "right") %>% 
  select(-Confidence) %>% 
  as.matrix()

counts_formatted = counts %>% as.matrix()

ps = phyloseq(sample_data(parkinsons_metadata_filt_cat),
              otu_table(counts_formatted, taxa_are_rows = T),
              tax_table(taxonomy_formatted))
```


## Alpha rarefaction curve
```{r alpha_rarefaction}
# Alpha rarefaction curve
rarecurve(t(data.frame(ps@otu_table)),
          step = 1000)

# Based on the alpha rarefaction curve, a sampling depth of 6000 will be used because at this sampling depth, all of the samples appear to reach the plateau.

# Rarefy the data based on the rarefaction depth chosen
set.seed(421)
psrare = ps %>% rarefy_even_depth(sample.size = 6000, rngseed = 421)
```

## Diversity Analysis (using DAI categorized PD cohort)

### Beta carotene

-   Alpha Diversity

```{r alpha_diversity_analysis_for_beta_carotene}
set.seed(421)
p = plot_richness(psrare, x = "Tert_BC",
                  measures = c("Observed", "Shannon", "Chao1", "Simpson"),
                  color = "Tert_BC")
p

# Modify the graph
# Pull the data from the plot:
pdata = p$data

str(pdata) # variable denotes metric, value denotes value

# Graph the plot again
p_formatted = pdata %>% ggplot(aes(Tert_BC, 
                                   value, 
                                   fill = Tert_BC)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2, size = 0.03) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  ylab("Alpha Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(fill = "Beta Carotene Intake Categorization")
p_formatted

# Run the Wilcoxon test
comparisons_BC = list(c("BC Low","BC High"),
                      c("BC Low", "BC Medium"),
                      c("BC Medium", "BC High"))

p_formatted + 
  stat_compare_means(comparisons = comparisons_BC, method = "wilcox.test", step.increase = 0.4, size = 6) +
  scale_y_continuous(expand = expansion(mult = c(0, .4))) #c(bottom, top)

```
-   Beta Diversity
```{r beta_diversity_analysis_for_beta_carotene}
# Calculate the beta-curtis distance between each sample pair
ps_bray = phyloseq::distance(psrare, method = "bray")
View(as.matrix(ps_bray)) # Each pairwise sample included

# MDS scaling
set.seed(421)
mds = metaMDS(ps_bray)

# Extract data
mds_data = mds$points %>% as.data.frame %>% 
  merge(sample_data(psrare), by = "row.names", sort = F)

# Plot it
plot = mds_data %>% 
  ggplot(aes(MDS1, MDS2, color = Tert_BC)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size = 18) +
  labs(color = "Beta Carotene DAI Categorization")
plot

# Run a PERMANOVA test on Tert
metadata_perm = sample_data(psrare) %>% data.frame()

# Single variable, which is Tert
stats_univar = adonis2(ps_bray ~ Tert_BC, data = metadata_perm)

# Two variables - adding Sex as a covariable to see if controlling for sex makes a difference
stats_multivar = adonis2(ps_bray ~ Tert_BC + Sex + Age + entacapone + Energy_kcal, data = metadata_perm, by = 'margin') 

# Combine into a table
stats = bind_rows("Univariate" = stats_univar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  "Multivariate" = stats_multivar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  .id = "Model") # First column is Model, either Uni or Multi

# Add some formatting
stats = stats %>% rename(Pval = `Pr(>F)`) %>% 
  filter(!is.na(Pval)) # Remove the irrelevant lines

# Select groups of interest for comparison
to_compare_one = c("BC Low", "BC High")

# List samples that belong to these groups of interest
samples_to_keep_one = metadata_perm %>% filter(Tert_BC %in% to_compare_one) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_one = as.matrix(ps_bray)[samples_to_keep_one, samples_to_keep_one] %>% as.dist()

# Run a PERMANOVA 
stats_univar_one = adonis2(ps_bray_sub_one ~ Tert_BC,
                       data = metadata_perm %>% filter(Tert_BC %in% to_compare_one))
stats_univar_one # p-value is 0.298
```

```{r Combine All Dietary Antioxidants Together for Alpha Diversity Analysis}

# Combines all the dietary antioxidant intake together and use the Dietary Antioxidant Index (DAI) for analysis, which is another method mentioned in the paper.

# Convert to data frame
sam_df_analysis_2 = data.frame(ps@sam_data)


# Put the dataset back into polyseq object
sample_data(ps) = sample_data(sam_df_analysis_2_DAI)

# Rarefy the data based on the rarefaction depth chosen
psrare = ps %>% rarefy_even_depth(sample.size = 4000, rngseed = 421)

# Calculate alpha diversity 
set.seed(421)
p = plot_richness(psrare, x = "Tert",
                  measures = c("Observed", "Shannon", "Chao1", "Simpson"),
                  color = "composite_DAI")
p

# Modify the graph
# Pull the data from the plot:
pdata = p$data

str(pdata) # variable denotes metric, value denotes value

# Graph the plot again
p_formatted = pdata %>% ggplot(aes(Tert, 
                                   value, 
                                   fill = Tert)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2, size = 0.03) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  ylab("Alpha Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(fill = "Composite DAI Categorization")
p_formatted

# Run the Wilcoxon test
comparisons_composite_DAI = list(c("Low","High"),
                                 c("Low", "Medium"),
                                 c("Medium", "High"))

p_formatted + 
  stat_compare_means(comparisons = comparisons_composite_DAI, method = "wilcox.test", step.increase = 0.4, size = 6) +
  scale_y_continuous(expand = expansion(mult = c(0, .4))) #c(bottom, top)

```

-   **Calculate the differences in Bray–Curtis beta diversity between groups with different levels of dietary antioxidant intake (DAI)**

```{r Combine All Dietary Antioxidants Together for Beta Diversity Analysis}
# Rarefy the data based on the rarefaction depth chosen
psrare = ps %>% rarefy_even_depth(sample.size = 4000, rngseed = 421)

# Calculate the beta-curtis distance between each sample pair
ps_bray = phyloseq::distance(psrare, method = "bray")
View(as.matrix(ps_bray)) # Each pairwise sample included

# MDS scaling
set.seed(421)
mds = metaMDS(ps_bray)

# Extract data
mds_data = mds$points %>% as.data.frame %>% 
  merge(sample_data(psrare), by = "row.names", sort = F)

# Plot it
plot = mds_data %>% 
  ggplot(aes(MDS1, MDS2, color = Tert)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size = 18) +
  labs(color = "DAI Categorization")
plot

# Run a PERMANOVA test on Tert
metadata_perm = sample_data(psrare) %>% data.frame()

# Single variable, which is Tert
stats_univar = adonis2(ps_bray ~ Tert, data = metadata_perm)

# Two variables - adding Sex as a covariable to see if controlling for sex makes a difference
stats_multivar = adonis2(ps_bray ~ Tert + Sex, data = metadata_perm, by = 'margin') 

# Combine into a table
stats = bind_rows("Univariate" = stats_univar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  "Multivariate" = stats_multivar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  .id = "Model") # First column is Model, either Uni or Multi

# Add some formatting
stats = stats %>% rename(Pval = `Pr(>F)`) %>% 
  filter(!is.na(Pval)) # Remove the irrelevant lines
```

```{r Compare between low and medium DAI groups}
# Select groups of interest for comparison - compare low DAI with medium DAI first
to_compare_one = c("Low", "Medium")

# List samples that belong to these groups of interest
samples_to_keep_one = metadata_perm %>% filter(Tert %in% to_compare_one) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_one = as.matrix(ps_bray)[samples_to_keep_one, samples_to_keep_one] %>% as.dist()

# Run a PERMANOVA 
stats_univar_one = adonis2(ps_bray_sub_one ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_one))
stats_univar_one # p-value is 0.744
```

```{r Compare between low and high DAI groups}
# Select groups of interest for comparison - compare low DAI with high DAI next
to_compare_two = c("Low", "High")

# List samples that belong to these groups of interest
samples_to_keep_two = metadata_perm %>% filter(Tert %in% to_compare_two) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_two = as.matrix(ps_bray)[samples_to_keep_two, samples_to_keep_two] %>% as.dist()

# Run a PERMANOVA 
stats_univar_two = adonis2(ps_bray_sub_two ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_two))
stats_univar_two #p-value is 0.05
```

```{r Compare between medium and high DAI groups}
# Select groups of interest for comparison - compare medium DAI with high DAI next
to_compare_three = c("Medium", "High")

# List samples that belong to these groups of interest
samples_to_keep_three = metadata_perm %>% filter(Tert %in% to_compare_three) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_three = as.matrix(ps_bray)[samples_to_keep_three, samples_to_keep_three] %>% as.dist()

# Run a PERMANOVA 
stats_univar_three = adonis2(ps_bray_sub_two ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_three))
stats_univar_three #p-value is 0.251

```

# PART 3: Taxonomic Analysis

1.  **What indicator taxa are associated with each of your groups of interest?**

    1.  Notes:

        1.  Run this at the Genus level

        2.  Remove any genera with a mean relative abundance of less than 0.0001 beforehand

        3.  Filter the results using a stat cutoff of \>0.7 and a P value cutoff of \<0.05.

        4.  Use dim() to print the dimensions of the filtered data frame below the code chunk

```{r}
# Aggregate ASVs to the Genus level
ps_genus = tax_glom(ps, "Genus")

# Convert phyloseq to relatvie abundance
ps_relab = transform(ps_genus, "compositional")

# Apply abundance filter to keep taxa that are at least 0.1% abundant on average
ps_filt = filter_taxa(ps_relab, function(x) mean(x) > 0.0001, TRUE)

otu_table = data.frame(otu_table(ps_filt))

set.seed(421)



# Beta carotene
# Use multipatt to cluster samples by group using a vector
indval_BC = multipatt(t(otu_table),
                      cluster = ps_filt@sam_data$Tert_BC,
                      control = how(nperm = 999))

# Use the summary() function to view the ISA results
summary(indval_BC, indvalcomp = TRUE)

# Save the statistics to a data frame
indval_table_BC = as.data.frame(indval_BC$sign)

# Filter the results using a stat cutoff of >0.7 and a P value cutoff of <0.05
filtered_table_BC = indval_table_BC %>% 
  filter(stat > 0.7 & p.value < 0.05) %>% 
  rownames()

dim(filtered_table_BC) # prints NULL

```

2.  **What taxa are differentially abundant in each of our groups of interest?**

    1.  Notes:

        1.  Run this at the Genus level
        2.  Remove any genera with a mean relative abundance of less than 0.0001 beforehand
        3.  Set prv_cut to 0.1 and p_adj_method to 'BH'
        4.  Filter the data frame to only include rows where the relevant diff_robust column is TRUE.
        5.  Use dim() to print the dimensions of the filtered data frame below the code chunk

```{r}
# Tax_glom
ps_glom = tax_glom(ps, "Genus")

# Convert phyloseq to relatvie abundance
ps_relab_two = transform(ps_glom, "compositional")

# Apply abundance filter to keep taxa that are at least 0.1% abundant on average
ps_filt_two = filter_taxa(ps_relab_two, function(x) mean(x) > 0.0001, TRUE)

otu_table = data.frame(otu_table(ps_filt_two))

set.seed(421)
out = ancombc2(data = ps_glom,
               fix_formula = "Tert_BC",
               p_adj_method = "BH",
               prv_cut = 0.05)

statistical_table = out$res

# Filter stats to only include taxa that are differentially abundant between low and medium DAI groups
taxa_to_plot_one = statistical_table %>% 
  filter(diff_robust_Tert_BCBCMedium == T) # 0 variable

# Filter stats to only include taxa that are differentially abundant between low and high DAI groups
taxa_to_plot_two = statistical_table %>% 
  filter(diff_robust_TertHigh == T) # 0 variablef

```
