---
title: "Diversity (alpha and beta) and Taxa Analysis"
output: html_document
---

```{r libraries, include = FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("phyloseq")
BiocManager::install("microbiome")
BiocManager::install("ANCOMBC")

library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ANCOMBC)
```

## Load the data

```{r load_data}
library(readxl)
parkinsons_metadata = read_excel('Datasets/parkinsons_metadata.xlsx') %>% 
  column_to_rownames("SampleID")# Just like assignment 4
taxonomy = read.delim('Datasets/taxonomy.tsv', row.names = 1) # from qiime2
counts = read.delim('Datasets/feature-table.txt', skip = 1, row.names = 1)  # from qiime2
```

# PART 1: Make a Phyloseq object

Using the data we loaded above and code from Module 12, create a phyloseq object called ps.

### Marks = 1 pt

```{r}
# Write your code here
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon,
           into = c("Domain", "Phylum", "Class", "Order", "Family",
                    "Genus", "Species"),
           sep = ";", fill = "right") %>% 
  select(-Confidence) %>% 
  as.matrix()

counts_formatted = counts %>% as.matrix()

ps = phyloseq(sample_data(parkinsons_metadata),
              otu_table(counts_formatted, taxa_are_rows = T),
              tax_table(taxonomy_formatted))
```

# PART 2: Diversity

### Marks = 2.5

**Using code from Module 13, do ONE of the following.**

-   **Option 1: Calculate the difference in Shannon alpha diversity between your groups of interest**

    -   Notes:

        -   Use rarecurve() to justify a reasonable sampling depth (print plot and write a short comment to justify)

            -   Don't worry about maximizing \# features in this case, just base your answer on the curve. A range will be accepted

        -   Calculate the metric as seen in the module

        -   Generate an appropriate publication-ready plot

        -   Add a Wilcoxon test directly to the plot using stat_compare_means(). Adjust the text size so that it is reasonably easy to read.

```{r To Prepare for Alpha Diversity Analysis}
# Convert to data frame
sam_df_analysis_1 = data.frame(ps@sam_data)

# Because we want to explore the impact of varying levels of dietary antioxidant (excluding irons) intake on the gut microbiome composition and structure of PD patients, we choose to only look at patients that have the full set of dietary information. Therefore, individuals with missing information on dietary antioxidant intake will be excluded from the metadata prior to downstream analyses. And control groups will also be excluded so that we are only doing analyses using PD patients. 
sam_df_PD_cohort_with_NA_removed = sam_df_analysis_1 %>% 
  filter(Disease == "PD") %>% 
  filter(if_all(c(Alpha_carotene, Beta_carotene, Vitamin_A, Vitamin_C, Vitamin_E), ~ !is.na(.)))
```

```{r Vitamin Analysis}
# To analyze how varying levels of various vitamin intake changes the alpha diversity of the gut microbiome in PD patients, we compare each of the five dietary antioxidant intake to their respective recommended dietary allowance (RDA) for Canadians established by Health Canada. Based on the paper, the PD cohort was categorized into two groups: individuals with vitamin intakes less than 2/3 of the RDA and those with intakes greater than or equal to 2/3 of the RDA.

# RDA values for vitamins reference: https://www.canada.ca/en/health-canada/services/food-nutrition/healthy-eating/dietary-reference-intakes/tables/reference-values-vitamins.html#tbl1

# Vitamin C: RDA is 75 mg/day for females and 90 mg/day for males
# Categorize vitamin C intake (mg) based on reference RDA
sam_df_PD_cohort_with_NA_removed = sam_df_PD_cohort_with_NA_removed %>% 
  mutate(Vitamin_C_categorized = ifelse((Sex == "Female" & Vitamin_C < 2/3 * 75) | (Sex == "Male" & Vitamin_C < 2/3 * 90),
                                        "low-intake VC",
                                        "high-intake VC")) %>% 
  relocate(Vitamin_C_categorized, .after = Vitamin_C)

# To see how many individuals in the PD cohort have low- and high-intakes of Vitamin C respectively
sam_df_PD_cohort_with_NA_removed %>% 
  count(Vitamin_C_categorized == "low-intake VC")
# In the PD cohort, 164 individuals have high-intake of vitamin C, 20 individuals have low-intake of vitamin C.


# Vitamin A: RDA is 700 ug/day for females and 900 ug/day for males
# Categorize vitamin A intake (mcg) based on reference RDA
sam_df_PD_cohort_with_NA_removed = sam_df_PD_cohort_with_NA_removed %>% 
  mutate(Vitamin_A_categorized = ifelse((Sex == "Female" & Vitamin_A < 2/3 * 700) | (Sex == "Male" & Vitamin_A < 2/3 * 900),
                                        "low-intake VA",
                                        "high-intake VA")) %>% 
  relocate(Vitamin_A_categorized, .after = Vitamin_A)

# To see how many individuals in the PD cohort have low- and high-intakes of Vitamin A respectively
sam_df_PD_cohort_with_NA_removed %>% 
  count(Vitamin_A_categorized == "low-intake VA")
# In the PD cohort, 38 inviduals have high-intake of vitamin A, 146 individuals have low-intake of vitamin A. 



# Vitamin E: 15 mg/day for both males and females
# Categorize vitamin E intake (mg) based on reference RDA
sam_df_PD_cohort_with_NA_removed = sam_df_PD_cohort_with_NA_removed %>% 
  mutate(Vitamin_E_categorized = ifelse(Vitamin_E < 2/3 * 15,
                                        "low-intake VE",
                                        "high-intake VE")) %>% 
  relocate(Vitamin_E_categorized, .after = Vitamin_E)

# To see how many individuals in the PD corhort have low- and high-intakes of Vitamin E respectively
sam_df_PD_cohort_with_NA_removed %>% 
  count(Vitamin_E_categorized == "low-intake VE")
# In the PD cohort, 114 individuals have high-intake of vitamin E, 70 individuals have low-intake of vitamin E.



  # Put the dataset back into polyseq object
sample_data(ps) = sample_data(sam_df_PD_cohort_with_NA_removed)

# Alpha rarefaction curve
rarecurve(t(data.frame(ps@otu_table)),
          step = 1000)

# Based on the alpha rarefaction curve, a sampling depth of 4000 will be used because at this sampling depth, all of the samples appear to reach the plateau.

# Rarefy the data based on the rarefaction depth chosen
psrare = ps %>% rarefy_even_depth(sample.size = 4000, rngseed = 421)



# Calculate alpha diversity for vitamin C
set.seed(421)
p = plot_richness(psrare, x = "Vitamin_C_categorized",
                  measures = c("Observed", "Shannon", "Chao1", "Simpson"),
                  color = "Vitamin_C_categorized")
p

# Modify the graph
# Pull the data from the plot:
pdata = p$data

str(pdata) # variable denotes metric, value denotes value

# Graph the plot again
p_formatted = pdata %>% ggplot(aes(Vitamin_C_categorized, 
                                   value, 
                                   fill = Vitamin_C_categorized)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2, size = 0.03) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  ylab("Alpha Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(fill = "Vitamin C Intake Categorization")
p_formatted

# Run the Wilcoxon test
comparisons_vitamin_C = list(c("low-intake VC", "high-intake VC"))

p_formatted + 
  stat_compare_means(comparisons = comparisons_vitamin_C, method = "wilcox.test", size = 6) +
  scale_y_continuous(expand = expansion(mult = c(0, .4))) #c(bottom, top)


# Calculate alpha diversity for vitamin A
set.seed(421)
p = plot_richness(psrare, x = "Vitamin_A_categorized",
                  measures = c("Observed", "Shannon", "Chao1", "Simpson"),
                  color = "Vitamin_A_categorized")
p

# Modify the graph
# Pull the data from the plot:
pdata = p$data

str(pdata) # variable denotes metric, value denotes value

# Graph the plot again
p_formatted = pdata %>% ggplot(aes(Vitamin_A_categorized, 
                                   value, 
                                   fill = Vitamin_A_categorized)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2, size = 0.03) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  ylab("Alpha Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(fill = "Vitamin A Intake Categorization")
p_formatted

# Run the Wilcoxon test
comparisons_vitamin_A = list(c("low-intake VA", "high-intake VA"))

p_formatted + 
  stat_compare_means(comparisons = comparisons_vitamin_A, method = "wilcox.test", size = 6) +
  scale_y_continuous(expand = expansion(mult = c(0, .4))) #c(bottom, top)

```

```{r Combine All Dietary Antioxidants Together for Analysis}

# Combines all the dietary antioxidant intake together and use the Dietary Antioxidant Index (DAI) for analysis, which is another method mentioned in the paper.

# Convert to data frame
sam_df_analysis_2 = data.frame(ps@sam_data)

# Calculate the mean intake and corresponding standard deviation for each of the five dietary antioxidants across the entire PD cohort. To prepare for downstream analysis, all dietary antioxidant intakes are standardized to mg to ensure consistency in units. 
mean_VA = mean(sam_df_analysis_2$Vitamin_A/1000)
sd_VA = sd(sam_df_analysis_2$Vitamin_A/1000)

mean_VC = mean(sam_df_analysis_2$Vitamin_C)
sd_VC = sd(sam_df_analysis_2$Vitamin_C)

mean_VE = mean(sam_df_analysis_2$Vitamin_E)
sd_VE = sd(sam_df_analysis_2$Vitamin_E)

mean_AC = mean(sam_df_analysis_2$Alpha_carotene/1000)
sd_AC = sd(sam_df_analysis_2$Alpha_carotene/1000)

mean_BC = mean(sam_df_analysis_2$Beta_carotene/1000)
sd_BC = sd(sam_df_analysis_2$Beta_carotene/1000)

# Assign a DAI to each patient in the PD cohort
sam_df_analysis_2_DAI = sam_df_analysis_2 %>% 
  mutate(composite_DAI = (Vitamin_A/1000 - mean_VA)/sd_VA + (Vitamin_C - mean_VC)/sd_VC + (Vitamin_E - mean_VE)/sd_VE + (Alpha_carotene/1000 - mean_AC)/sd_AC + (Beta_carotene/1000 - mean_BC)/sd_BC)

# Print the last column of the sam_df_analysis_2_DAI data frame to better visualize it
sam_df_analysis_2_DAI %>% select(last_col())

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert = quantile(sam_df_analysis_2_DAI$composite_DAI, c(0:3/3))
# Then, classify the values
sam_df_analysis_2_DAI$Tert = with(sam_df_analysis_2_DAI,
                                  cut(composite_DAI,
                                      vTert,
                                      include.lowest = T,
                                      labels = c("Low", "Medium", "High")))
# Print the Tert column to better visualize it
sam_df_analysis_2_DAI %>% select(last_col())



# Put the dataset back into polyseq object
sample_data(ps) = sample_data(sam_df_analysis_2_DAI)

# Rarefy the data based on the rarefaction depth chosen
psrare = ps %>% rarefy_even_depth(sample.size = 4000, rngseed = 421)

# Calculate alpha diversity 
set.seed(421)
p = plot_richness(psrare, x = "Tert",
                  measures = c("Observed", "Shannon", "Chao1", "Simpson"),
                  color = "composite_DAI")
p

# Modify the graph
# Pull the data from the plot:
pdata = p$data

str(pdata) # variable denotes metric, value denotes value

# Graph the plot again
p_formatted = pdata %>% ggplot(aes(Tert, 
                                   value, 
                                   fill = Tert)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.2, size = 0.03) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable, ncol = 2, scales = "free_y") +
  ylab("Alpha Diversity") + xlab(NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(fill = "Composite DAI Categorization")
p_formatted

# Run the Wilcoxon test
comparisons_composite_DAI = list(c("Low","High"),
                                 c("Low", "Medium"),
                                 c("Medium", "High"))

p_formatted + 
  stat_compare_means(comparisons = comparisons_composite_DAI, method = "wilcox.test", step.increase = 0.4, size = 6) +
  scale_y_continuous(expand = expansion(mult = c(0, .4))) #c(bottom, top)

```

-   **Option 2: Calculate the difference in Bray-Curtis beta diversity between wild type and susceptible mice**

    -   Notes:

        -   Use rarecurve() to justify a reasonable sampling depth (print plot and write a short comment to justify)

            -   Don't worry about maximizing \# features in this case, just base your answer on the curve. A range will be accepted

        -   Calculate the metric as seen in the module

        -   Generate an appropriate publication-ready plot. Don't forget to include ellipses.

        -   Run a PERMANOVA. You don't need to add it to the plot - just write your code so that the statistical output is printed below your code chunk.

```{r Combine All Dietary Antioxidants Together for Analysis}
# Rarefy the data based on the rarefaction depth chosen
psrare = ps %>% rarefy_even_depth(sample.size = 4000, rngseed = 421)

# Calculate the beta-curtis distance between each sample pair
ps_bray = phyloseq::distance(psrare, method = "bray")
View(as.matrix(ps_bray)) # Each pairwise sample included

# MDS scaling
set.seed(421)
mds = metaMDS(ps_bray)

# Extract data
mds_data = mds$points %>% as.data.frame %>% 
  merge(sample_data(psrare), by = "row.names", sort = F)

# Plot it
plot = mds_data %>% 
  ggplot(aes(MDS1, MDS2, color = Tert)) +
  geom_point() +
  stat_ellipse() +
  theme_classic(base_size = 18) +
  labs(color = "DAI Categorization")
plot

# Run a PERMANOVA test on Tert
metadata_perm = sample_data(psrare) %>% data.frame()

# Single variable, which is Tert
stats_univar = adonis2(ps_bray ~ Tert, data = metadata_perm)

# Two variables - adding Sex as a covariable to see if controlling for sex makes a difference
stats_multivar = adonis2(ps_bray ~ Tert + Sex, data = metadata_perm, by = 'margin') 

# Combine into a table
stats = bind_rows("Univariate" = stats_univar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  "Multivariate" = stats_multivar %>% as.data.frame() %>% 
                    rownames_to_column ("Variable"),
                  .id = "Model") # First column is Model, either Uni or Multi

# Add some formatting
stats = stats %>% rename(Pval = `Pr(>F)`) %>% 
  filter(!is.na(Pval)) # Remove the irrelevant lines
```

```{r Compare between low and medium DAI groups}
# Select groups of interest for comparison - compare low DAI with medium DAI first
to_compare_one = c("Low", "Medium")

# List samples that belong to these groups of interest
samples_to_keep_one = metadata_perm %>% filter(Tert %in% to_compare_one) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_one = as.matrix(ps_bray)[samples_to_keep_one, samples_to_keep_one] %>% as.dist()

# Run a PERMANOVA 
stats_univar_one = adonis2(ps_bray_sub_one ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_one))
stats_univar_one # p-value is 0.744
```

```{r Compare between low and high DAI groups}
# Select groups of interest for comparison - compare low DAI with high DAI next
to_compare_two = c("Low", "High")

# List samples that belong to these groups of interest
samples_to_keep_two = metadata_perm %>% filter(Tert %in% to_compare_two) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_two = as.matrix(ps_bray)[samples_to_keep_two, samples_to_keep_two] %>% as.dist()

# Run a PERMANOVA 
stats_univar_two = adonis2(ps_bray_sub_two ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_two))
stats_univar_two #p-value is 0.05
```

```{r Compare between medium and high DAI groups}
# Select groups of interest for comparison - compare medium DAI with high DAI next
to_compare_three = c("Medium", "High")

# List samples that belong to these groups of interest
samples_to_keep_three = metadata_perm %>% filter(Tert %in% to_compare_three) %>% 
  rownames()

# Subset the distance matrix to only include those samples
ps_bray_sub_three = as.matrix(ps_bray)[samples_to_keep_three, samples_to_keep_three] %>% as.dist()

# Run a PERMANOVA 
stats_univar_three = adonis2(ps_bray_sub_two ~ Tert,
                       data = metadata_perm %>% filter(Tert %in% to_compare_three))
stats_univar_three #p-value is 0.251

```

# PART 3: Taxonomic Analysis

### Marks = 1.5

**Run ONE of the following analyses:**

1.  **What indicator taxa are associated with each of your groups of interest?**

    1.  Notes:

        1.  Run this at the Genus level

        2.  Remove any genera with a mean relative abundance of less than 0.0001 beforehand

        3.  Filter the results using a stat cutoff of \>0.7 and a P value cutoff of \<0.05.

        4.  Use dim() to print the dimensions of the filtered data frame below the code chunk

```{r}
# Aggregate ASVs to the Genus level
ps_genus = tax_glom(ps, "Genus")

# Convert phyloseq to relatvie abundance
ps_relab = transform(ps_genus, "compositional")

# Apply abundance filter to keep taxa that are at least 0.1% abundant on average
ps_filt = filter_taxa(ps_relab, function(x) mean(x) > 0.0001, TRUE)

otu_table = data.frame(otu_table(ps_filt))

set.seed(421)



# Vitamin A
# Use multipatt to cluster samples by group using a vector
indval_VA = multipatt(t(otu_table),
                      cluster = ps_filt@sam_data$Vitamin_A_categorized,
                      control = how(nperm = 999))

# Use the summary() function to view the ISA results
summary(indval_VA, indvalcomp = TRUE)

# Save the statistics to a data frame
indval_table_VA = as.data.frame(indval_VA$sign)

# Filter the results using a stat cutoff of >0.7 and a P value cutoff of <0.05
filtered_table_VA = indval_table_VA %>% 
  filter(stat > 0.7 & p.value < 0.05) %>% 
  rownames()

dim(filtered_table_VA) # prints NULL


  
# Vitamin C
# Use multipatt to cluster samples by group using a vector
indval_VC = multipatt(t(otu_table),
                      cluster = ps_filt@sam_data$Vitamin_C_categorized,
                      control = how(nperm = 999))

# Use the summary() function to view the ISA results
summary(indval_VC, indvalcomp = TRUE)

# Save the statistics to a data frame
indval_table_VC = as.data.frame(indval_VC$sign)

# Filter the results using a stat cutoff of >0.7 and a P value cutoff of <0.05
filtered_table_VC = indval_table_VC %>% 
  filter(stat > 0.7 & p.value < 0.05) %>% 
  rownames()

dim(filtered_table_VC) # prints NULL



# Combine all the dietary antioxidants together
# Use multipatt to cluster samples by group using a vector
indval_combined = multipatt(t(otu_table),
                      cluster = ps_filt@sam_data$Tert,
                      control = how(nperm = 999))

# Use the summary() function to view the ISA results
summary(indval_combined, indvalcomp = TRUE)

# Save the statistics to a data frame
indval_table_combined = as.data.frame(indval_combined$sign)

# Filter the results using a stat cutoff of >0.7 and a P value cutoff of <0.05
filtered_table_combined = indval_table_combined %>% 
  filter(stat > 0.7 & p.value < 0.05) %>% 
  rownames()

dim(filtered_table_combined) # prints NULL

```

2.  **What taxa are differentially abundant in each of your groups of interest?**

    1.  Notes:

        1.  Run this at the Genus level
        2.  Remove any genera with a mean relative abundance of less than 0.0001 beforehand
        3.  Set prv_cut to 0.1 and p_adj_method to 'BH'
        4.  Filter the data frame to only include rows where the relevant diff_robust column is TRUE.
        5.  Use dim() to print the dimensions of the filtered data frame below the code chunk
```{r}
# Tax_glom
ps_glom = tax_glom(ps, "Genus")

# Convert phyloseq to relatvie abundance
ps_relab_two = transform(ps_glom, "compositional")

# Apply abundance filter to keep taxa that are at least 0.1% abundant on average
ps_filt_two = filter_taxa(ps_relab_two, function(x) mean(x) > 0.0001, TRUE)

otu_table = data.frame(otu_table(ps_filt_two))

set.seed(421)
out = ancombc2(data = ps_glom,
               fix_formula = "Tert",
               p_adj_method = "BH",
               prv_cut = 0.05)

statistical_table = out$res

# Filter stats to only include taxa that are differentially abundant between low and medium DAI groups
taxa_to_plot_one = statistical_table %>% 
  filter(diff_robust_TertMedium == T) # 0 variable

# Filter stats to only include taxa that are differentially abundant between low and high DAI groups
taxa_to_plot_two = statistical_table %>% 
  filter(diff_robust_TertHigh == T) # 0 variablef

```
