---
title: "Beta-Carotene Picrust differential abundance"
author: "Cynthia"
date: "2026-03-14"
output: html_document
---

# Introductions
For aim 3 we wanted to identify the functional pathways of significance. This script is designed to identify and visualize functional differences in the gut microbiome between individuals with Low vs. High Beta-Carotene(BC) intake within a Parkinson’s Disease cohort. It utilizes PICRUst2 pathway predictions and applies Maaslin2 for robust differential abundance testing. Pathway plot, Heatmap plot, PCA plot, and volcano plot code is also included here.

## Environment Setup and Data Loading
This section starts the analysis by loading required bioinformatic and visualization libraries and preparing the metadata.
```{r libraries, include=FALSE}
# Core data manipulation and plotting
library(tidyverse)
library(phyloseq)
library(ggrepel)
library(stringr)

# analysis specific tools
library(ggpicrust2)
library(Maaslin2)

# visualization-specific tools
library(pheatmap)
library(ggExtra)

# Loading metadata and filter for specific BC comparison groups
metadata = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# focusing only on low vs high BC tertiles and drop unused factor levels
metadata_sub = metadata %>% filter(Tert_BC %in% c("BC Low", "BC High")) %>%
  mutate(Tert_BC = fct_drop(as.factor(Tert_BC)))
```

## Differential Abundance Analysis (Maaslin2)
We perform the core statistical testing here, using Maaslin2 to find pathways that differ significantly while adjusting for potential confounders like Sex, Age, and entacapone use.
```{r}
# Load predicted MetaCyc pathway abundances from QIIME2/PICRUst2 output
pathway_abundance = read.delim('../../../QIIME2_outputs/path_abun_unstrat.tsv',row.names = 1)

# Subset the abundance table to match the samples present in our filtered metadata
path_filt = pathway_abundance %>% select(all_of(rownames(metadata_sub)))

# Run Differential Abundance Analysis (DAA) using Maaslin2
# Adjusting for Sex, Age, and Entacapone usage as fixed effects
diff_abun = pathway_daa(abundance = path_filt,
                   metadata = metadata_sub,
                   group = "Tert_BC",
                   daa_method = "Maaslin2",
                   fixed_effects = c("Sex", "Age", "entacapone"),
                   select = NULL,
                   reference = "BC Low")

# Annotate pathway codes (MetaCyc IDs) with human-readable descriptions
daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = diff_abun,
                                              ko_to_kegg = FALSE)
 
# Cache the results to avoid re-running intensive computations
saveRDS(daa_annotated_results_df,'daa_annotated_results_df.rds')

# reading saved results
readRDS('daa_annotated_results_df.rds')

```
## Manual Log2 Fold Change Calculation
Maaslin2 provided coefficients that appeared very out of range of what we expected so we did manual $log_2$ fold change (L2FC) calculations for each. This section calculates the magnitude of change between groups.
```{r}
# STEP 1: Convert your counts to relative abundance
path_relab = apply(path_filt, MARGIN = 2, function(x) x/sum(x))

# STEP 2: make a data table with all your samples, counts per pathway, and variable of interest.
path_long = path_relab %>% 
  as.data.frame() %>% # apply() turned it into a matrix, need a data frame
  rownames_to_column('pathway') %>% # move the function names to a column
  pivot_longer(cols = -pathway, names_to = 'sample_name', values_to = 'value')

# adding my variable of interest
combined_table = path_long %>% left_join(metadata_sub %>% rownames_to_column('sample_name'), by = 'sample_name')
head(combined_table)

# STEP 3: Add Pseudocount
# half of the smallest non-zero value in the relative abundance data:
pseudocount_value = min(combined_table$value[combined_table$value>0])/2
pseudocount_value

# Then add it to all the abundances:
combined_table$value = combined_table$value + pseudocount_value
min(combined_table$value) # Now equal to the pseudocount


# STEP 4: for every pathway, calculate the average relative abundance for each group of interest.
avg_abundances = combined_table %>% 
  group_by(pathway,Tert_BC) %>% 
  summarize(value = mean(value)) %>% 
  ungroup() 

# STEP 5: Take the ratio of your groups of interest, then apply a log2 fold change
ratio = avg_abundances %>% 
  pivot_wider(names_from = Tert_BC, values_from = value) %>% 
  mutate(ratio = `BC High`/`BC Low`) %>% 
  mutate(log2fc = log2(ratio)) 

head (ratio)

# STEP 6: Filter for absolute log2FC >=1
ratio_filt = ratio %>% 
  filter(abs(log2fc)>=1)

# Filter the differential abundance results to only include those above the log2 fold change cutoff
daa_results_log2fc_filtered = daa_annotated_results_df %>% 
  filter(feature %in% ratio_filt$pathway)

# Extracting final significant pathway names for downstream use
significant_pathways = daa_results_log2fc_filtered %>% 
  filter(p_adjust < 0.05) %>%
  pull(feature)

# save
saveRDS(significant_pathways,'BC_significant_pathways.rds')
```

## Visualizing Significant Pathways (Error Bar Plots)
These plots visualize the mean proportion of pathways in each group and their associated effect sizes. The final plot is in our manuscript as Fig. S4
```{r}
# Load custom formatting function for pathway plots
source('../ggpicrust2_errorbar_function_fixed.R')

# filter for significant pathways for plotting 
significant_pathways_df = daa_results_log2fc_filtered %>% 
  filter(p_adjust < 0.05)

# find NA descriptions
view(significant_pathways_df)

# manually filling in missing description with info from MetaCyc website
significant_pathways_df$description[17] <- "Entner-Doudoroff pathway I"

# plot function
sanity_peb = pathway_errorbar_fixed(abundance = path_filt, 
                     daa_results_df = significant_pathways_df, 
                     Group = metadata_sub$Tert_BC, 
                     wrap_label = F, wraplength=60,
                     fc_cutoff = 0, order_by_log = F,
                     p_values_threshold = 0.05, 
                     ko_to_kegg = F, 
                     p_value_bar = TRUE, 
                     x_lab = "description")

# Save the final sanity checked version with the tags
ggsave("../../../Results/Figures/Aim_3/Beta-carotene/BC-picrust-sanity-final-with-labels.png", plot = sanity_peb, width = 12, height = 7, dpi = 300)
```

## Top 10 Magnitude Changes
This identifies and plots the pathways with the greatest magnitude(not just significant) shifts between Vitamin C levels. In our manuscript this is Fig.3D
```{r}
# filter for top 10 greatest log2fc
top_10 <- ratio %>%
  filter(pathway %in% significant_pathways_df$feature)%>%
  arrange(desc(abs(log2fc))) %>%  # Sort by greatest magnitude L2FC
  head(10)

# dataframe for plotting
daa_results_top10 = daa_annotated_results_df %>% 
  filter(feature %in% top_10$pathway)

# plotting only those pathways
peb10 = pathway_errorbar_fixed(abundance = path_filt, 
                     daa_results_df = daa_results_top10, 
                     Group = metadata_sub$Tert_BC, 
                     wrap_label = F, wraplength=60,
                     fc_cutoff = 0, order_by_log = F,
                     p_values_threshold = 1, 
                     ko_to_kegg = F, 
                     p_value_bar = TRUE, 
                     x_lab = "description")

# save
ggsave("../../../Results/Figures/Aim_3/Beta-carotene/BC-picrust-sig-pathways10.png",
       plot = peb10, width = 18, height = 15, dpi = 300)
```


## Correlation Heatmap
This heatmap explores how the significant pathways correlate with one another across samples, revealing functional modules. This was not used directly in our manuscript, only when characterizing initial trends observed at beginning of analysis.
```{r identical features}
# pulling features out
sig_features = daa_results_log2fc_filtered %>% 
  filter(p_adjust < 0.05, abs(log2_fold_change) > 0) %>% 
  pull('feature') %>% unique()

# Normalize counts to relative abundance (Total Sum Scaling) to ensure 
# correlations are based on biological proportions rather than sequencing depth
path_relab = path_filt %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
colSums(path_relab) 

# Calculate Pearson correlation between significant pathways
stats = path_relab %>% t() %>%  
  as.data.frame() %>% 
  select(all_of(sig_features)) %>% 
  cor(method = 'pearson') #Pearson tests between every feature pair

# Define the color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(40)

# Define breaks for the color scale
breaks <- seq(-1, 1, length.out = 41)  # 40 colors + 1 for the endpoint

# Map technical IDs to readable descriptions for plot labels
name_lookup <- daa_results_log2fc_filtered %>%
  filter(feature %in% sig_features) %>%
  select(feature, description) %>% 
  distinct()

readable_names <- name_lookup$description[match(rownames(stats), name_lookup$feature)]

# manually fixing NA descriptions
readable_names[17] <- "Entner-Doudoroff pathway I"

# wrapping labels
rownames(stats) <- str_wrap(rownames(stats), width = 35)
colnames(stats) <- str_wrap(colnames(stats), width = 35)

view(readable_names)

# plotting and trying to minimize white space
pheatmap(stats, 
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",          
         color = color_palette, 
         breaks = breaks,
         border_color = "white",
         display_numbers = TRUE, 
         number_format = "%.2f",      
         fontsize_number = 6,          # shrinks the numbers in cells
         fontsize_row = 8,
         fontsize_col = 8,
         angle_col = 315,               
         treeheight_row = 15,         
         treeheight_col = 15, 
         legend = TRUE,                
         main = NA,                    
         filename = "../../../Results/Figures/Aim_3/Beta-carotene/MetaCyc-BC_Heatmap_Final-new.png",
         width = 15,                   
         height = 13)

```

## PCA and Volcano Visualization
This final section provides a global overview of sample similarity (PCA) and a summary of all statistical tests (Volcano Plot). The volcano plot generated here was used in our manuscript as Fig.3C
### PCA (beta diversity)
```{r PCA}
# Define custom color palettes for the High and Low groups to ensure 
# consistency across all manuscript figures.
line_cols <- c(
  "BC High" = "#D96B6B",
  "BC Low"  = "#4F97D7"
)


fill_cols <- c(
  "BC High" = "#E07B7B",
  "BC Low"  = "#6FA8DC"
)

# Generate the base PCA plot to visualize functional (pathway) similarity.
# Marginal curves are disabled here because they will be added manually for better control.
pca_plot <- pathway_pca(
  abundance = path_filt,
  metadata = metadata_sub, 
  group = "Tert_BC",
  show_marginal = FALSE
) +
  labs(color = "Tert_BC", fill = "Tert_BC") +
  scale_color_manual(values = line_cols) +
  scale_fill_manual(values = fill_cols) +
  theme_classic() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA)
  )

# Use ggMarginal to add density plots to the top and right margins. 
# This helps visualize where the bulk of samples in each group fall along PC1 and PC2.
final_plot <- ggMarginal(
  pca_plot,
  type = "density",
  groupFill = TRUE,
  groupColour = TRUE,
  alpha = 0.35
)

# save
ggsave(
  "../../../Results/Figures/Aim_3/Beta-carotene/BC-pathway-pca.png",
  plot = final_plot,
  width = 8,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

### Volcano plot
```{r}
# Join the p-values from Maaslin2 with the log2FC calculated
plot_data <- daa_annotated_results_df %>%
  left_join(ratio, by = c("feature" = "pathway")) %>%
  mutate(
    # Define significance categories for coloring
    status = case_when(
      p_adjust < 0.05 & log2fc >= 1 ~ "More Abundant (BC Low)",
      p_adjust < 0.05 & log2fc <= -1 ~ "Less Abundant (BC High)",
      TRUE ~ "Not Significant"
    )
  )

# Create the Volcano Plot
volcano_plot <- ggplot(plot_data, aes(x = log2fc, y = -log10(p_adjust), color = status)) +
  geom_point(alpha = 0.6, size = 2) +
  # Add threshold lines
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  # Custom colors
  scale_color_manual(values = c("More Abundant (BC Low)" = "#D55E00", 
                                "Less Abundant (BC High)" = "#0072B2", 
                                "Not Significant" = "grey70")) +
  geom_text_repel(
    data = subset(plot_data, p_adjust < 0.01 & abs(log2fc) > 1.5),
    aes(label = description), 
    size = 3,
    max.overlaps = 10,
    show.legend = FALSE
  ) +
  theme_minimal() +
  labs(
    x = expression(log[2]~"Fold Change"),
    y = expression(-log[10]~"Adjusted P-value"),
    color = "Significance"
  ) +
  theme(legend.position = "bottom")

# generating volcano plot
volcano_plot <- volcano_plot +
  theme(
    text = element_text(size = 16), 
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold")
  )

# Save
ggsave("../../../Results/Figures/Aim_3/Beta-carotene/BC-volcano-plot.png", 
       plot = volcano_plot, width = 10, height = 8, dpi = 300)
```

