---
title: "Data Filtering and Wrangling Preparation for Diversity and Taxa Analysis"
output: html_document
date: "2026-03-05"
---

## Library loading

```{r libraries, include = FALSE}
library(tidyverse)
```

## Load the dataset

```{r load_data}
parkinsons_metadata = read.delim('../Datasets/parkinsons_metadata.txt')
```

## Data Filtering

Because we want to explore the impact of varying levels of dietary antioxidant (excluding irons) intake on the gut microbiome composition and structure of PD patients, we choose to only look at patients that have the full set of dietary information. Therefore, individuals with missing information on dietary antioxidant intake will be excluded from the metadata prior to downstream analyses. And control groups will also be excluded so that we are only doing analyses using PD patients.

```{r data_filtering}
parkinsons_metadata_PD_cohort_NA_removed = parkinsons_metadata %>% 
  filter(Disease == "PD") %>% 
  filter(if_all(c(Alpha_carotene, Beta_carotene, Vitamin_A, Vitamin_C, Vitamin_E), ~ !is.na(.)))
```

## Dietary Antioxidant Related Calculations

To analyze how varying levels of various vitamin intake changes the alpha diversity of the gut microbiome in PD patients, we will be categorizing PD patients into groups based on their daily dietary antioxidant intake values using two different classification approaches.

### Method 1:

Compare each of the five dietary antioxidant intake to their respective recommended dietary allowance (RDA) for Canadians established by Health Canada. Based on the paper, the PD cohort was categorized into two groups: individuals with vitamin intakes less than 2/3 of the RDA and those with intakes greater than or equal to 2/3 of the RDA.

-   RDA values for vitamins reference:
    <https://www.canada.ca/en/health-canada/services/food-nutrition/healthy-eating/dietary-reference-intakes/tables/reference-values-vitamins.html#tbl1>)

-   Note: there is no official RDA for alpha and beta carotene

#### Vitamin A

```{r vitamin_A}
# Vitamin A: RDA is 700 ug/day for females and 900 ug/day for males
# Categorize vitamin A intake (mcg) based on reference RDA
parkinsons_metadata_filt_cat = parkinsons_metadata_PD_cohort_NA_removed %>% 
  mutate(Vitamin_A_cat = ifelse((Sex == "Female" & Vitamin_A < 2/3 * 700) | (Sex == "Male" & Vitamin_A < 2/3 * 900),
                                "low-intake VA",
                                "high-intake VA")) %>% 
  relocate(Vitamin_A_cat, .after = Vitamin_A)
# To see how many individuals in the PD cohort have low- and high-intakes of Vitamin A respectively
parkinsons_metadata_filt_cat %>% 
  count(Vitamin_A_cat == "low-intake VA")
# In the PD cohort, 38 inviduals have high-intake of vitamin A, 146 individuals have low-intake of vitamin A. 
```

#### Vitamin C

```{r vitamin_C}
# Vitamin C: RDA is 75 mg/day for females and 90 mg/day for males
# Categorize vitamin C intake (mg) based on reference RDA
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Vitamin_C_cat = ifelse((Sex == "Female" & Vitamin_C < 2/3 * 75) | (Sex == "Male" & Vitamin_C < 2/3 * 90),
                                "low-intake VC",
                                "high-intake VC")) %>% 
  relocate(Vitamin_C_cat, .after = Vitamin_C)

# To see how many individuals in the PD cohort have low- and high-intakes of Vitamin C respectively
parkinsons_metadata_filt_cat %>% 
  count(Vitamin_C_cat == "low-intake VC")
# In the PD cohort, 164 individuals have high-intake of vitamin C, 20 individuals have low-intake of vitamin C.
```

#### Vitamin E

```{r vitamin_C}
# Vitamin E: 15 mg/day for both males and females
# Categorize vitamin E intake (mg) based on reference RDA
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Vitamin_E_cat = ifelse(Vitamin_E < 2/3 * 15,
                                "low-intake VE",
                                "high-intake VE")) %>% 
  relocate(Vitamin_E_cat, .after = Vitamin_E)

# To see how many individuals in the PD corhort have low- and high-intakes of Vitamin E respectively
parkinsons_metadata_filt_cat %>% 
  count(Vitamin_E_cat == "low-intake VE")
# In the PD cohort, 114 individuals have high-intake of vitamin E, 70 individuals have low-intake of vitamin E.
```

### Method 2:

Use the Dietary Antioxidant Index (DAI) for analysis, which removes the need to use the RDA as a reference value. Therefore, when using this categorization method, we are able to combine all the dietary antioxidant intake together for analysis.

#### Vitamin A

```{r vitamin_A}
# Calculate the mean intake and corresponding standard deviation for Vitamin A across the entire PD cohort
mean_VA = mean(parkinsons_metadata_filt_cat$Vitamin_A)
sd_VA = sd(parkinsons_metadata_filt_cat$Vitamin_A)

# Assign a DAI score to each PD patient
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Vitamin_A_DAI = (Vitamin_A - mean_VA)/sd_VA) %>% 
  relocate(Vitamin_A_DAI, .after = Vitamin_A_cat)

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_VA = quantile(parkinsons_metadata_filt_cat$Vitamin_A_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_VA = with(parkinsons_metadata_filt_cat,
                                            cut(Vitamin_A_DAI,
                                                vTert_VA,
                                                include.lowest = T,
                                                labels = c("VA Low", "VA Medium", "VA High"))) 

# Relocate the Tert_VA column so that it is to the right of Vitamin_A_DAI
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  relocate(Tert_VA, .after = Vitamin_A_DAI)
```

#### Vitamin C

```{r vitamin_C}
mean_VC = mean(parkinsons_metadata_filt_cat$Vitamin_C)
sd_VC = sd(parkinsons_metadata_filt_cat$Vitamin_C)

# Assign a DAI score to each PD patient
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Vitamin_C_DAI = (Vitamin_C - mean_VC)/sd_VC) %>% 
  relocate(Vitamin_C_DAI, .after = Vitamin_C_cat)

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_VC = quantile(parkinsons_metadata_filt_cat$Vitamin_C_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_VC = with(parkinsons_metadata_filt_cat,
                                            cut(Vitamin_C_DAI,
                                                vTert_VC,
                                                include.lowest = T,
                                                labels = c("VC Low", "VC Medium", "VC High"))) 

# Relocate the Tert_VC column so that it is to the right of Vitamin_C_DAI
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  relocate(Tert_VC, .after = Vitamin_C_DAI)
```

#### Vitamin E

```{r vitamin_E}
mean_VE = mean(parkinsons_metadata_filt_cat$Vitamin_E)
sd_VE = sd(parkinsons_metadata_filt_cat$Vitamin_E)

# Assign a DAI score to each PD patient
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Vitamin_E_DAI = (Vitamin_E - mean_VE)/sd_VE) %>% 
  relocate(Vitamin_E_DAI, .after = Vitamin_E_cat)

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_VE = quantile(parkinsons_metadata_filt_cat$Vitamin_E_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_VE = with(parkinsons_metadata_filt_cat,
                                            cut(Vitamin_E_DAI,
                                                vTert_VE,
                                                include.lowest = T,
                                                labels = c("VE Low", "VE Medium", "VE High"))) 

# Relocate the Tert_VE column so that it is to the right of Vitamin_E_DAI
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  relocate(Tert_VE, .after = Vitamin_E_DAI)
```

#### Alpha Carotene

```{r alpha_carotene}
mean_AC = mean(parkinsons_metadata_filt_cat$Alpha_carotene)
sd_AC = sd(parkinsons_metadata_filt_cat$Alpha_carotene)

# Assign a DAI score to each PD patient
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Alpha_carotene_DAI = (Alpha_carotene - mean_AC)/sd_AC) %>% 
  relocate(Alpha_carotene_DAI, .after = Alpha_carotene)

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_AC = quantile(parkinsons_metadata_filt_cat$Alpha_carotene_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_AC = with(parkinsons_metadata_filt_cat,
                                            cut(Alpha_carotene_DAI,
                                                vTert_AC,
                                                include.lowest = T,
                                                labels = c("AC Low", "AC Medium", "AC High"))) 

# Relocate the Tert_AC column so that it is to the right of Alpha_carotene_DAI
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  relocate(Tert_AC, .after = Alpha_carotene_DAI)
```

#### Beta Carotene

```{r beta_carotene}
mean_BC = mean(parkinsons_metadata_filt_cat$Beta_carotene)
sd_BC = sd(parkinsons_metadata_filt_cat$Beta_carotene)

# Assign a DAI score to each PD patient
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(Beta_carotene_DAI = (Beta_carotene - mean_BC)/sd_BC) %>% 
  relocate(Beta_carotene_DAI, .after = Beta_carotene)

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_BC = quantile(parkinsons_metadata_filt_cat$Beta_carotene_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_BC = with(parkinsons_metadata_filt_cat,
                                            cut(Beta_carotene_DAI,
                                                vTert_BC,
                                                include.lowest = T,
                                                labels = c("BC Low", "BC Medium", "BC High"))) 

# Relocate the Tert_BC column so that it is to the right of Beta_carotene_DAI
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  relocate(Tert_BC, .after = Beta_carotene_DAI)
```

#### Combine all five dietart antioxidants together

```{r dietary_antioxidant_combined}
# Calculate the mean intake and corresponding standard deviation for each of the five dietary antioxidants across the entire PD cohort. To prepare for downstream analysis, all dietary antioxidant intakes are standardized to mg to ensure consistency in units. 
mean_VA_combine = mean(parkinsons_metadata_filt_cat$Vitamin_A/1000)
sd_VA_combine = sd(parkinsons_metadata_filt_cat$Vitamin_A/1000)

mean_VC_combine = mean(parkinsons_metadata_filt_cat$Vitamin_C)
sd_VC_combine = sd(parkinsons_metadata_filt_cat$Vitamin_C)

mean_VE_combine = mean(parkinsons_metadata_filt_cat$Vitamin_E)
sd_VE_combine = sd(parkinsons_metadata_filt_cat$Vitamin_E)

mean_AC_combine = mean(parkinsons_metadata_filt_cat$Alpha_carotene/1000)
sd_AC_combine = sd(parkinsons_metadata_filt_cat$Alpha_carotene/1000)

mean_BC_combine = mean(parkinsons_metadata_filt_cat$Beta_carotene/1000)
sd_BC_combine = sd(parkinsons_metadata_filt_cat$Beta_carotene/1000)

# Assign a DAI to each patient in the PD cohort
parkinsons_metadata_filt_cat = parkinsons_metadata_filt_cat %>% 
  mutate(composite_DAI = (Vitamin_A/1000 - mean_VA_combine)/sd_VA_combine + (Vitamin_C - mean_VC_combine)/sd_VC_combine + (Vitamin_E - mean_VE_combine)/sd_VE_combine + (Alpha_carotene/1000 - mean_AC_combine)/sd_AC_combine + (Beta_carotene/1000 - mean_BC_combine)/sd_BC_combine)

# Print the last column of the parkinsons_metadata_filt_cat data frame to better visualize it
parkinsons_metadata_filt_cat %>% select(last_col())

# Assign categorical groups to all PD patients by splitting them into tertiles, which means all the PD patients are divided into three even groups.
# First, find the tertiles
vTert_combine = quantile(parkinsons_metadata_filt_cat$composite_DAI, c(0:3/3))
# Then, classify the values
parkinsons_metadata_filt_cat$Tert_combine = with(parkinsons_metadata_filt_cat,
                                                 cut(composite_DAI,
                                                     vTert_combine,
                                                     include.lowest = T,
                                                     labels = c("Combine Low", "Combine Medium", "Combine High")))
# Print the Tert_combine column to better visualize it
parkinsons_metadata_filt_cat %>% select(last_col())
```

## Export the modified dataset for downstream analyses

```{r dataset export}
write.csv(parkinsons_metadata_filt_cat, file = "../Datasets/parkinsons_metadata_filt_cat.csv", row.names = FALSE)
```
