---
title: "Indicator Species Analysis (final version)"
authors: Carrie Sun, Yanting Xiong
output: html_document
date: "2026-03-20"
---

## Introduction
This R Markdown script performs an Indicator Species Analysis (ISA) to identify specific bacterial genera that are significantly associated with either High or Low intake levels of various dietary antioxidants in Parkinson's Disease patients. By comparing these extreme groups, the analysis identifies potential "biomarker" taxa that thrive or diminish in response to specific nutrient environments.

## Library loading
This section initializes the analysis environment with packages for microbiome data management and ecological association testing.
```{r library_loading}
library(tidyverse)
library(phyloseq)
library(indicspecies)
```

## Dataset loading
The purpose of this section is to import raw data and create filtered metadata subsets that focus strictly on the comparison between the Low (Tertile 1) and High (Tertile 3) intake groups.
```{r dataset_loading}
# Load in the filtered and wrangled Parkinson's metadata
parkinsons_metadata_filt_cat = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# For each variable prefixed with "Tert_", the metadata will be filtered to only include PD patients in the low and high intake categories, and the resulting subsets will be used for downstream analyses
parkinsons_metadata_filt_cat_AC_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_AC == "AC Low" | Tert_AC == "AC High")

parkinsons_metadata_filt_cat_BC_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_BC == "BC Low" | Tert_BC == "BC High")

parkinsons_metadata_filt_cat_VA_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_VA == "VA Low" | Tert_VA == "VA High")

parkinsons_metadata_filt_cat_VC_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_VC == "VC Low" | Tert_VC == "VC High")

parkinsons_metadata_filt_cat_VE_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_VE == "VE Low" | Tert_VE == "VE High")

parkinsons_metadata_filt_cat_combine_LH_only = parkinsons_metadata_filt_cat %>% 
  filter(Tert_combine == "Combine Low" | Tert_combine == "Combine High")

# Load in feature table exported from QIIME2
counts = read.delim('../../../QIIME2_outputs/feature-table.txt', skip = 1, row.names = 1)

# Load in taxonomy table exported from QIIME2
taxonomy = read.delim('../../../QIIME2_outputs/taxonomy.tsv', row.names = 1)
```

## Create a Phyloseq Object
This chunk integrates counts, taxonomy, and the filtered metadata into unified objects for each nutrient comparison path.
```{r phyloseq_object}
# Taxonomy
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon, 
           into = c("Domain","Phylum","Class","Order",
                    "Family","Genus","Species"),
           sep = ";", fill = "right") %>% 
  select(-Confidence) %>% 
  as.matrix()

# Counts
counts_formatted = counts %>% as.matrix()

# Create the phyloseq object for each dietary antioxidant 
ps_AC = phyloseq(sample_data(parkinsons_metadata_filt_cat_AC_LH_only),
                 otu_table(counts_formatted, taxa_are_rows = T),
                 tax_table(taxonomy_formatted))

ps_BC = phyloseq(sample_data(parkinsons_metadata_filt_cat_BC_LH_only),
                 otu_table(counts_formatted, taxa_are_rows = T),
                 tax_table(taxonomy_formatted))

ps_VA = phyloseq(sample_data(parkinsons_metadata_filt_cat_VA_LH_only),
                 otu_table(counts_formatted, taxa_are_rows = T),
                 tax_table(taxonomy_formatted))

ps_VC = phyloseq(sample_data(parkinsons_metadata_filt_cat_VC_LH_only),
                 otu_table(counts_formatted, taxa_are_rows = T),
                 tax_table(taxonomy_formatted))

ps_VE = phyloseq(sample_data(parkinsons_metadata_filt_cat_VE_LH_only),
                 otu_table(counts_formatted, taxa_are_rows = T),
                 tax_table(taxonomy_formatted))

ps_combine = phyloseq(sample_data(parkinsons_metadata_filt_cat_combine_LH_only),
                      otu_table(counts_formatted, taxa_are_rows = T),
                      tax_table(taxonomy_formatted))

```

## Preprocessing
This section simplifies the data by aggregating it to the genus level, converting it to relative abundance, and removing very rare taxa to reduce statistical noise.
```{r preprocessing}
# Aggregate ASVs to higher taxonomic level (genus level in this case)
ps_genus_AC = tax_glom(ps_AC,"Genus")

ps_genus_BC = tax_glom(ps_BC,"Genus")

ps_genus_VA = tax_glom(ps_VA,"Genus")

ps_genus_VC = tax_glom(ps_VC,"Genus")

ps_genus_VE = tax_glom(ps_VE,"Genus")

ps_genus_combine = tax_glom(ps_combine,"Genus")

# Convert phyloseq to relative abundance
ps_relab_AC = microbiome::transform(ps_genus_AC, "compositional")

ps_relab_BC = microbiome::transform(ps_genus_BC, "compositional")

ps_relab_VA = microbiome::transform(ps_genus_VA, "compositional")

ps_relab_VC = microbiome::transform(ps_genus_VC, "compositional")

ps_relab_VE = microbiome::transform(ps_genus_VE, "compositional")

ps_relab_combine = microbiome::transform(ps_genus_combine, "compositional")

# Filter to only include taxa that have a mean abundance of greater than or equal to 0.01%
ps_filt_AC = filter_taxa(ps_relab_AC, function(x) mean(x) >= 0.0001, TRUE)

ps_filt_BC = filter_taxa(ps_relab_BC, function(x) mean(x) >= 0.0001, TRUE)

ps_filt_VA = filter_taxa(ps_relab_VA, function(x) mean(x) >= 0.0001, TRUE)

ps_filt_VC = filter_taxa(ps_relab_VC, function(x) mean(x) >= 0.0001, TRUE)

ps_filt_VE = filter_taxa(ps_relab_VE, function(x) mean(x) >= 0.0001, TRUE)

ps_filt_combine = filter_taxa(ps_relab_combine, function(x) mean(x) >= 0.0001, TRUE)

# Extract the otu_table as a data frame because only the otu_table, instead of the phyloseq object, will be used in indicator taxa analysis
otu_table_AC = data.frame(otu_table(ps_filt_AC))

otu_table_BC = data.frame(otu_table(ps_filt_BC))

otu_table_VA = data.frame(otu_table(ps_filt_VA))

otu_table_VC = data.frame(otu_table(ps_filt_VC))

otu_table_VE = data.frame(otu_table(ps_filt_VE))

otu_table_combine = data.frame(otu_table(ps_filt_combine))
```

## Indicator Analysis
This section performs the Multi-level Pattern Analysis for each nutrient. The multipatt function identifies which taxa are uniquely associated with specific intake groups.

## Alpha Carotene

```{r alpha_carotene}
set.seed(421)
indval_AC = multipatt(t(otu_table_AC),
                      cluster = ps_filt_AC@sam_data$Tert_AC,
                      control = how(nperm = 999))

summary(indval_AC, indvalcomp = TRUE) 
# No indicator taxa associated with the high intake group
# No indicator taxa associated with the low intake group

# Extract into a table
indval_table_AC = as.data.frame(indval_AC$sign)

```

## Beta Carotene

```{r beta_carotene}
set.seed(421)
indval_BC = multipatt(t(otu_table_BC),
                      cluster = ps_filt_BC@sam_data$Tert_BC,
                      control = how(nperm = 999))

summary(indval_BC, indvalcomp = TRUE) 
# Two indicator taxa associated with the high intake group
  # 630532cabb74720c1204579e6fe741b4 (p-value = 0.045, stat = 0.403)
  # a030e5d5ff9c75a027e6cc58ff02cf7c (p-value = 0.026, stat = 0.370)

# One indicator taxa associated with the low intake group
  # 4c8288bfbd76958c0c094d87b97650f8 (p-value = 0.006, stat = 0.715)

# Extract into a table
indval_table_BC = as.data.frame(indval_BC$sign)

```

## Vitamin A

```{r vitamin_a}
set.seed(421)
indval_VA = multipatt(t(otu_table_VA),
                      cluster = ps_filt_VA@sam_data$Tert_VA,
                      control = how(nperm = 999))

summary(indval_VA, indvalcomp = TRUE) 
# Two indicator taxa associated with the high intake group
  # 7bf7ea59cac793e65c0bc293eabf44dc (p-value = 0.045, stat = 0.388)
  # b287c87a16ef231b6b93625eceae36b1 (p-value = 0.013, stat = 0.313)

# Extract into a table
indval_table_VA = as.data.frame(indval_VA$sign)

```

## Vitamin C

```{r vitamin_c}
set.seed(421)
indval_VC = multipatt(t(otu_table_VC),
                      cluster = ps_filt_VC@sam_data$Tert_VC,
                      control = how(nperm = 999))

summary(indval_VC, indvalcomp = TRUE) 
# Three indicator taxa associated with the high intake group
  # a3d1f4355f204d4d49ad2bea3b801265 (p-value = 0.043, stat = 0.411)
  # 6ca3f1e7d3011097ec41585e61307d26 (p-value = 0.015, stat = 0.361) 
  # a033d13f0f20b5a9847c0d750f9b257b (p-value = 0.044, stat = 0.305)

# Three indicator taxa associated with the low intake group
  # 4c8288bfbd76958c0c094d87b97650f8 (p-value = 0.003, stat = 0.736)
  # 7bf7ea59cac793e65c0bc293eabf44dc (p-value = 0.022, stat = 0.436)
  # f0135ea5c9854954ff851185c476e91c (p-value = 0.005, stat = 0.379)

# Extract into a table
indval_table_VC = as.data.frame(indval_VC$sign)

```

## Vitamin E

```{r vitamin_e}
set.seed(421)
indval_VE = multipatt(t(otu_table_VE),
                      cluster = ps_filt_VE@sam_data$Tert_VE,
                      control = how(nperm = 999))

summary(indval_VE, indvalcomp = TRUE) 
# Two indicator taxa associated with the high intake group
  # b7fb7b8de1b4e013e7255cec428893a7 (p-value = 0.036, stat = 0.354)
  # c97980fcfebe112b1e72f25624597e10 (p-value = 0.047, stat = 0.256)

# One indicator taxa associated with the low intake group
  # 93f6e99ff381954b541d29cc24c2be7c (p-value = 0.036, stat = 0.424)

# Extract into a table
indval_table_VE = as.data.frame(indval_VE$sign)

```

## Combined

```{r combined}
set.seed(421)
indval_combine = multipatt(t(otu_table_combine),
                           cluster = ps_filt_combine@sam_data$Tert_combine,
                           control = how(nperm = 999))

summary(indval_combine, indvalcomp = TRUE) 
# Two indicator taxa associated with the high intake group
  # 4661cd41d2df6a4bf70276894e6cc779 (p-value = 0.047, stat = 0.404)
  # 8408059b62784f6a2740d42ac2aa1069 (p-value = 0.017, stat = 0.347)

# Extract into a table
indval_table_combine = as.data.frame(indval_combine$sign)

```
