---
title: "Picrust_differential_abundance"
author: "Cynthia"
date: "2026-03-14"
output: html_document
---

# Introductions
For aim 3 we wanted to identify the functional pathways of significance. This script is designed to identify and visualize functional differences in the gut microbiome between individuals with Low vs. High combined intake within a Parkinson’s Disease cohort. It utilizes PICRUst2 pathway predictions and applies Maaslin2 for robust differential abundance testing. Pathway plot, Heatmap plot, and PCA plot is also included here. Note that because the results indicated differences in combined groups were insignificant therefore manual log2fold change checks and volcano plots were not produced. Figures produced here are not included in final manuscript.

## Environment Setup and Data Loading
This section starts the analysis by loading required bioinformatic and visualization libraries and preparing the metadata. 
```{r libraries, include=FALSE}
# Core data manipulation and plotting
library(tidyverse)
library(stringr)

# analysis specific tools
library(ggpicrust2)
library(Maaslin2)

# visualization-specific tools
library(pheatmap)

# Loading metadata and filter for specific combined comparison groups
metadata = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# focusing only on low vs high combined tertiles and drop unused factor levels
metadata_sub = metadata %>% filter(Tert_combine %in% c("Combine Low", "Combine High")) %>%
  mutate(Tert_combine = fct_drop(as.factor(Tert_combine)))
```

## Differential Abundance Analysis (Maaslin2)
We perform the core statistical testing here, using Maaslin2 to find pathways that differ significantly while adjusting for potential confounders like Sex, Age, and entacapone use.
```{r}
# Load predicted MetaCyc pathway abundances from QIIME2/PICRUst2 output
pathway_abundance = read.delim('../../../QIIME2_outputs/path_abun_unstrat.tsv',row.names = 1)

# Subset the abundance table to match the samples present in our filtered metadata
path_filt = pathway_abundance %>% select(all_of(rownames(metadata_sub)))

# Run Differential Abundance Analysis (DAA) using Maaslin2
# Adjusting for Sex, Age, and Entacapone usage as fixed effects
diff_abun = pathway_daa(abundance = path_filt,
                   metadata = metadata,
                   group = "Tert_combine",
                   daa_method = "Maaslin2",
                   fixed_effects = c("Sex", "Age", "entacapone"),
                   select = NULL,
                   reference = "Combine Low")

# Annotate pathway codes (MetaCyc IDs) with human-readable descriptions
daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = diff_abun,
                                              ko_to_kegg = FALSE)
 
# Cache the results to avoid re-running intensive computations
saveRDS(daa_annotated_results_df,'daa_annotated_results_df.rds')

# reading saved results
readRDS('daa_annotated_results_df.rds')

```

## Visualizing Pathways (Error Bar Plots)
These plots visualize the mean proportion of pathways in each group and their associated effect sizes. Note that pathways plotted here are insignificant and only presented as a representation of this group. Figures produced are not included in our final manuscript.
```{r}
# Load custom formatting function for pathway plots
source('../ggpicrust2_errorbar_function_fixed.R')

# Filter out NA in p-values and log2 fold change columns
daa_annotated_results_df_filt = daa_annotated_results_df %>%
  filter(!is.na(p_adjust), !is.na(log2_fold_change))

# Adjusted P-value = 10 hits for plotting (not significant)
daa_annotated_results_df_filt %>% 
  filter(p_adjust < 0.06) %>% 
  nrow()

# Log2 Fold Change values = 15 hits (sufficient for plot)
daa_annotated_results_df_filt %>% 
  filter(abs(log2_fold_change) >= 5e-3) %>% 
  nrow()

# Log2 Fold change values + adjusted p value = 6 hits for plot
daa_annotated_results_df_filt %>% filter(p_adjust< 0.06, abs(log2_fold_change) > 5e-3) %>% nrow()

# Dataframe for plotting
daa_strict <- daa_annotated_results_df_filt %>% 
  filter(p_adjust < 0.06, abs(log2_fold_change) > 5e-3)

# plot
peb = pathway_errorbar_fixed(abundance = path_filt, 
                     daa_results_df = daa_strict, 
                     Group = metadata_sub$Tert_combine, 
                     wrap_label = T, wraplength=60,
                     fc_cutoff = 5e-3, order_by_log = T,
                     p_values_threshold = 0.06, 
                     order = "name", 
                     ko_to_kegg = FALSE, 
                     p_value_bar = TRUE, 
                     x_lab = "description")

# save
ggsave("../../../Results/Figures/Aim_3 (Figure_3, Figure_S3, Figure_S4)/Combined/combine-picrust-pathways.png",
       plot = peb, width = 18, height = 10)
```

## Correlation Heatmap
This heatmap explores how the significant pathways correlate with one another across samples, revealing functional modules. This was not used directly in our manuscript, only when characterizing initial trends observed at beginning of analysis. Note that pathways plotted here are insignificant and only presented as a representation of this group.
```{r identical features}
# working with same pathways as above
sig_features = daa_annotated_results_df %>% 
  filter(p_adjust < 0.06, abs(log2_fold_change) > 5e-3) %>% 
  pull('feature') %>% unique()

# Normalize counts to relative abundance (Total Sum Scaling) to ensure 
# correlations are based on biological proportions rather than sequencing depth
path_relab = path_filt %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
colSums(path_relab) # Should be all 1

# Calculate Pearson correlation between significant pathways
stats = path_relab %>% t() %>%  
  as.data.frame() %>% 
  select(all_of(sig_features)) %>% 
  cor(method = 'pearson') #Pearson tests between every feature pair

# Define the color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(40)

# Define breaks for the color scale
breaks <- seq(-1, 1, length.out = 41)  # 40 colors + 1 for the endpoint

# Lookup Names
name_lookup <- daa_annotated_results_df %>%
  filter(feature %in% sig_features) %>%
  select(feature, description) %>%
  distinct()

readable_names <- name_lookup$description[match(rownames(stats), name_lookup$feature)]
rownames(stats) <- readable_names
colnames(stats) <- readable_names

# plot to try to minimize white space
pheatmap(stats, 
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",          
         color = color_palette, 
         breaks = breaks,
         border_color = "white",
         display_numbers = TRUE,
         number_format = "%.2f",      
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 315,              # rotated to minimize white space
         treeheight_row = 25,         
         treeheight_col = 25, 
         legend = TRUE,               
         main = NA,                   
         filename = "../../../Results/Figures/Aim_3 (Figure_3, Figure_S3, Figure_S4)/Combined/MetaCyc-combine_Heatmap_Final-new.png",
         width = 8,                   
         height = 6)

```

## PCA Visualization
This final section provides a global overview of sample similarity (PCA). Note that pathways plotted here are insignificant and only presented as a representation of this group. Figures produced are not included in our final manuscript.
```{r PCA}
# Generate pathway PCA plot
pca = pathway_pca(abundance = path_filt,
            metadata = metadata_sub, 
            group = "Tert_combine")

# save pca
ggsave("../../../Results/Figures/Aim_3 (Figure_3, Figure_S3, Figure_S4)/Combined/AC-pathway-pca.png", 
       plot = pca, 
       width = 8, 
       height = 6, 
       dpi = 300)
```
