---
title: "Aim 4 - Indicator Taxon and MoCA Score"
author: Carrie Sun
output: html_document
date: "2026-03-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This script investigates whether the relative abundance of the indicator genus Escherichia-Shigella (identified in previous aims) is a significant predictor of cognitive performance, measured by MoCA scores, in Parkinson’s Disease patients, while adjusting for clinical covariates like Age, Sex, and medication.

## Load Library
Initialize the environment with essential packages for microbiome data manipulation and tidy data processing.
```{r}
library(tidyverse)
library(phyloseq)
library(ggplot2)
```

# Load dataset
Import raw feature tables, taxonomy assignments, and clinical metadata.
```{r Load datasets}
# Load in the filtered and wrangled Parkinson's metadata
parkinsons_metadata_filt_cat = read.csv("../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# Load in feature table exported from QIIME2
counts = read.delim('../../QIIME2_outputs/feature-table.txt', skip = 1, row.names = 1)

# Load in taxonomy table exported from QIIME2
taxonomy = read.delim('../../QIIME2_outputs/taxonomy.tsv', row.names = 1)

```

# Create a Phyloseq Object
Consolidate counts, taxonomy, and metadata into a unified object to ensure sample-data alignment.
```{r phyloseq_object}
# Counts
counts_formatted = counts %>% as.matrix()

# Taxonomy
taxonomy_formatted = taxonomy %>% 
  separate(col = Taxon, 
           into = c("Domain", "Phylum", "Class", "Order",
                    "Family", "Genus", "Species"),
           sep = ";", fill = "right") %>% 
  select(-Confidence) %>% 
  as.matrix()

ps = phyloseq(sample_data(parkinsons_metadata_filt_cat),
              otu_table(counts_formatted, taxa_are_rows = T),
              tax_table(taxonomy_formatted))

```

## Indicator Taxon Processing
Isolate the target taxon, normalize its abundance, and apply a transformation to correct for skewness.
```{r Load the indicator taxa}
# Extract the indicator taxa obtained from Aim 2 
taxa_id <- c("4c8288bfbd76958c0c094d87b97650f8") # Corresponds to Escherichia-Shigella

# Aggregate ASVs to higher taxonomic level (genus level in this case)
ps_genus = tax_glom(ps, "Genus")

# Convert phyloseq to relative abundance
ps_relab = transform_sample_counts(ps_genus, function(x) x / sum(x))

# Extract the phyloseq otu_table
# Transpose is applied to ensure samples are rows for easier 
# data-frame extraction, as taxa_are_rows is TRUE by default in this object.
otu = as.data.frame(otu_table(ps_relab))

# Make samples = rows
if (taxa_are_rows(ps_relab)) {
  otu = t(otu)
}

otu = as.data.frame(otu)   

# Keep only the taxon of interest
#drop= false will ensure keep the extracted column as dataframe
sub = otu[, taxa_id, drop = FALSE]

sub = data.frame(
  SampleID = rownames(otu),
  rel_abund = otu[, taxa_id]
)

# Raw relative abundance histogram -- right skewed, not suitable for linear regression model
relative_abundance_distribution = ggplot(sub, aes(x = rel_abund)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  labs(x = "Relative Abundance of Escherichia-Shigella", y = "Number of PD Patients") +
  theme_classic(base_size = 18) +
  theme(axis.title = element_text(face = "bold")) 

# export image Fig. S5
ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Indicator_taxon/Relative_abundance_distribution.png", plot = relative_abundance_distribution, width = 10, height = 7, dpi = 300)

# The lowest non-zero relative abundance is 0.0001518834

# Log transform relative abundance
# A Log10 transformation is used because raw microbial abundance 
# is highly right-skewed. We add a pseudocount of half the minimum non-zero value
sub$log_rel_abund = log10(sub$rel_abund + 0.0001518834/2)

hist(sub$log_rel_abund,
     breaks = 30,
     main = "Distribution of Log10 Relative Abundance",
     xlab = "log10(Relative Abundance + 0.0001518834/2)",
     ylab = "Frequency")
```

## Metadata Integration
Join the processed microbial abundance data with clinical records for statistical modeling.

```{r Add to Metadata}
meta <- data.frame(sample_data(ps_relab))

meta$SampleID <- rownames(meta)

meta2 <- merge(meta, sub, by = "SampleID")
```

# Linear Regression Model
Test the association between the indicator taxon and MoCA scores while controlling for confounders.
```{r Predict the MoCA based on the indicator taxa}

model <- lm(moca_total ~ log_rel_abund + Age + Sex + entacapone, data = meta2)

summary(model)

# For log10-transformed relative abundance, estimate = -0.03384; p = 0.8769

# Age seems to be the predictor of MoCA score.

```

# Linear Regression Plot
Generate a scatter plot with a linear regression line to visualize the relationship for the final manuscript.
```{r Generate Plot for Aim 4}

linear_regression_plot = ggplot(meta2, aes(x = log_rel_abund, y = moca_total)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Log-10 Relative Abundance of Escherichia-Shigella",
       y = "MoCA Score of PD Patients") +
  theme_classic(base_size = 18) +
  theme(axis.title = element_text(face = "bold")) 
linear_regression_plot

# Export the plot
ggsave("../../Results/Figures/Aim_4 (Figure_2, Figure_S5)/Indicator_taxon/Association Between Log-10 Transformed Relative Abundance of Escherichia-Shigella and MoCA Score in PD Patients.png", plot = linear_regression_plot, width = 10, height = 7, dpi = 300)

```


