---
title: "Vitamin C Picrust differential abundance"
author: "Cynthia"
date: "2026-03-14"
output: html_document
---

# Introductions
For aim 3 we wanted to identify the functional pathways of significance. This script is designed to identify and visualize functional differences in the gut microbiome between individuals with Low vs. High Vitamin C intake within a Parkinson’s Disease cohort. It utilizes PICRUst2 pathway predictions and applies Maaslin2 for robust differential abundance testing. Pathway plot, Heatmap plot, PCA plot, and volcano plot code is also included here.

## Environment Setup and Data Loading
This section starts the analysis by loading required bioinformatic and visualization libraries and preparing the metadata.
```{r libraries, include=FALSE}
# Core data manipulation and plotting
library(tidyverse)
library(stringr)

# analysis specific tools
library(ggpicrust2)
library(Maaslin2)

# visualization-specific tools
library(pheatmap)
library(ggExtra)

# Loading metadata and filter for specific Vitamin C (VC) comparison groups
metadata = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# focusing only on low vs high VC tertiles and drop unused factor levels
metadata_sub = metadata %>% filter(Tert_VC %in% c("VC Low", "VC High")) %>%
  mutate(Tert_VC = fct_drop(as.factor(Tert_VC)))
```

## Differential Abundance Analysis (Maaslin2)
We perform the core statistical testing here, using Maaslin2 to find pathways that differ significantly while adjusting for potential confounders like Sex, Age, and entacapone use.
```{r}
# Load predicted MetaCyc pathway abundances from QIIME2/PICRUst2 output
pathway_abundance = read.delim('../../../QIIME2_outputs/path_abun_unstrat.tsv',row.names = 1)

# Subset the abundance table to match the samples present in our filtered metadata
path_filt = pathway_abundance %>% select(all_of(rownames(metadata_sub)))

# Run Differential Abundance Analysis (DAA) using Maaslin2
# Adjusting for Sex, Age, and Entacapone usage as fixed effects
diff_abun = pathway_daa(abundance = path_filt,
                   metadata = metadata,
                   group = "Tert_VC",
                   daa_method = "Maaslin2",
                   fixed_effects = c("Sex", "Age", "entacapone"),
                   select = NULL,
                   reference = "VC Low")

# Annotate pathway codes (MetaCyc IDs) with human-readable descriptions
daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = diff_abun,
                                              ko_to_kegg = FALSE)
 
# Cache the results to avoid re-running intensive computations
saveRDS(daa_annotated_results_df,'daa_annotated_results_df.rds')

# reading saved results
readRDS('daa_annotated_results_df.rds')

```

## Manual Log2 Fold Change Calculation
Maaslin2 provided coefficients that appeared very out of range of what we expected so we did manual $log_2$ fold change (L2FC) calculations for each. This section calculates the magnitude of change between groups.
```{r}
# STEP 1: Convert your counts to relative abundance
path_relab = apply(path_filt, MARGIN = 2, function(x) x/sum(x))

table(colSums(path_relab))

# STEP 2: make a data table with all your samples, counts per pathway, and variable of interest.
path_long = path_relab %>% 
  as.data.frame() %>% 
  rownames_to_column('pathway') %>% 
  pivot_longer(cols = -pathway, names_to = 'sample_name', values_to = 'value')

# adding variable of interest
combined_table = path_long %>% left_join(metadata_sub %>% rownames_to_column('sample_name'), by = 'sample_name')
head(combined_table)

# STEP 3: Add Pseudocount
# half of the smallest non-zero value in the relative abundance data:
pseudocount_value = min(combined_table$value[combined_table$value>0])/2

# Then add it to all the abundances:
combined_table$value = combined_table$value + pseudocount_value
min(combined_table$value) # equal to the pseudocount

# STEP 4: for every pathway, calculate the average relative abundance for each group of interest.
avg_abundances = combined_table %>% 
  group_by(pathway,Tert_VC) %>%
  summarize(value = mean(value)) %>% 
  ungroup() 

# STEP 5: Take the ratio of your groups of interest, then apply a log2 fold change
ratio = avg_abundances %>% 
  pivot_wider(names_from = Tert_VC, values_from = value) %>% 
  mutate(ratio = `VC High`/`VC Low`) %>% 
  mutate(log2fc = log2(ratio)) 

head (ratio)

# STEP 6: Filter for absolute log2FC >=1

ratio_filt = ratio %>% 
  filter(abs(log2fc)>=1)

# Filter the differential abundance results to only include those above the log2 fold change cutoff
daa_results_log2fc_filtered = daa_annotated_results_df %>% 
  filter(feature %in% ratio_filt$pathway)

# Extracting final significant pathway names for downstream use
significant_pathways = daa_results_log2fc_filtered %>% 
  filter(p_adjust < 0.05) %>%
  pull(feature)

# save
saveRDS(significant_pathways,'VC_significant_pathways.rds')
```

## Visualizing Significant Pathways (Error Bar Plots)
These plots visualize the mean proportion of pathways in each group and their associated effect sizes. We split the results into two parts due to size constraints and to ensure legibility. The final manually combined plot is in our manuscript as Fig.S3
```{r}
# Load custom formatting function for pathway plots
source('../ggpicrust2_errorbar_function_fixed.R')

# Filter out NA in p-values and log2 fold change columns
daa_annotated_results_df_filt = daa_results_log2fc_filtered %>%
  filter(!is.na(p_adjust), !is.na(log2_fold_change))

# filter for only significant pathways = 37 hits
signif_path = daa_results_log2fc_filtered %>% 
  filter(p_adjust < 0.05)

# subset all 37 significant pathways as 2 separate data for plot
daa_part1 <- signif_path[1:18, ]
daa_part2 <- signif_path[19:37, ] # missing names for some

# Manually adding missing descriptions from MetaCyc
daa_part2$description[10] <- "D-erythronate degradation II"
daa_part2$description[11] <- "L-threonate degradation"
daa_part2$description[12] <- "glycogen biosynthesis III (from α-maltose 1-phosphate)"
daa_part2$description[13] <- "Entner-Doudoroff pathway I"

# Generate Error Bar plots for all 37 significant pathways
peb_part1 = pathway_errorbar_fixed(abundance = path_filt,
                                   daa_results_df = daa_part1, 
                                   Group = metadata_sub$Tert_VC, 
                                   wrap_label = T, wraplength = 45,
                                   fc_cutoff = 0,
                                   order_by_log = T,
                                   p_values_threshold = 1, 
                                   ko_to_kegg = FALSE, 
                                   p_value_bar = TRUE, 
                                   x_lab = "description")

peb_part2 = pathway_errorbar_fixed(abundance = path_filt,
                                   daa_results_df = daa_part2, 
                                   Group = metadata_sub$Tert_VC, 
                                   wrap_label = T, wraplength = 45,
                                   fc_cutoff = 0,
                                   order_by_log = T,
                                   p_values_threshold = 1, 
                                   ko_to_kegg = FALSE, 
                                   p_value_bar = TRUE, 
                                   x_lab = "description")

# Save plots
ggsave("../../../Results/Figures/Aim_3/Vitamin_C/VC-picrust-part1.png",
       plot = peb_part1, width = 18, height = 12)

ggsave("../../../Results/Figures/Aim_3/Vitamin_C/VC-picrust-part2.png",
       plot = peb_part2, width = 18, height = 12)

```

## Top 10 Magnitude Changes
This identifies and plots the pathways with the greatest magnitude(not just significant) shifts between Vitamin C levels. In our manuscript this is Fig.3B.
```{r}
# filter for top 10 greatest magnitude changes
top_10 <- ratio %>%
  filter(pathway %in% signif_path$feature)%>%
  arrange(desc(abs(log2fc))) %>%  # Sort by greatest magnitude L2FC
  head(10)

daa_results_top10 = daa_annotated_results_df %>% 
  filter(feature %in% top_10$pathway)

# plotting only those pathways
peb10 = pathway_errorbar_fixed(abundance = path_filt, 
                     daa_results_df = daa_results_top10, 
                     Group = metadata_sub$Tert_VC, 
                     wrap_label = F, wraplength = 60,
                     fc_cutoff = 0, order_by_log = F,
                     p_values_threshold = 1, 
                     ko_to_kegg = F, 
                     p_value_bar = TRUE, 
                     x_lab = "description")

# save
ggsave("../../../Results/Figures/Aim_3/Vitamin_C/VC-picrust-sig-pathways10.png",
       plot = peb10, width = 18, height = 15, dpi = 300)
```

## Correlation Heatmap
This heatmap explores how the significant pathways correlate with one another across samples, revealing functional modules. This was not used directly in our manuscript, only when characterizing initial trends observed at beginning of analysis.
```{r identical features}
# pulling features out of signif_path
sig_features = signif_path %>% 
  filter(p_adjust < 1, abs(log2_fold_change) > 0) %>% 
  pull('feature') %>% unique()

# Normalize counts to relative abundance (Total Sum Scaling) to ensure 
# correlations are based on biological proportions rather than sequencing depth
path_relab = path_filt %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
colSums(path_relab)

# Calculate Pearson correlation between significant pathways
stats = path_relab %>% t() %>%  
  as.data.frame() %>% 
  select(all_of(sig_features)) %>% 
  cor(method = 'pearson') 

# Define the color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(40)

# Define breaks for the color scale
breaks <- seq(-1, 1, length.out = 41)  # 40 colors + 1 for the endpoint

# Map technical IDs to readable descriptions for plot labels
name_lookup <- daa_annotated_results_df %>%
  filter(feature %in% sig_features) %>%
  select(feature, description) %>% 
  distinct()

readable_names <- name_lookup$description[match(rownames(stats), name_lookup$feature)]

# manually finding missing names
readable_names[28] <- "D-erythronate degradation II"
readable_names[29] <- "L-threonate degradation"
readable_names[30] <- "glycogen biosynthesis III (from α-maltose 1-phosphate)"
readable_names[31] <- "Entner-Doudoroff pathway I"

rownames(stats) <- readable_names
colnames(stats) <- readable_names

# Wrap row names (pathway names) at 30 characters
rownames(stats) <- str_wrap(rownames(stats), width = 30)

# Wrap column names at 30 characters
colnames(stats) <- str_wrap(colnames(stats), width = 30)

# plotting and trying to minimize white space
pheatmap(stats, 
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",          
         color = color_palette, 
         breaks = breaks,
         border_color = "white",
         display_numbers = TRUE,
         number_format = "%.2f",      
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 315,              # rotated to minimize white space
         treeheight_row = 25,         
         treeheight_col = 25, 
         legend = TRUE,               
         main = NA,                   # Use NA to properly suppress the title space
         filename = "../../../Results/Figures/Aim_3/Vitamin_C/MetaCyc-VC_Heatmap_Final-new.png",
         width = 35,                   
         height = 33)

```

## PCA and Volcano Visualization
This final section provides a global overview of sample similarity (PCA) and a summary of all statistical tests (Volcano Plot). The volcano plot generated here was used in our manuscript as Fig.3A
### PCA (beta diversity)
```{r PCA}
# Generate the base PCA plot WITHOUT the marginals
pca_plot = pathway_pca(abundance = path_filt,
                        metadata = metadata_sub, 
                        group = "Tert_VC",
                        show_marginal = FALSE) # Disable the hardcoded curves

# labels
pca_plot = pca_plot + 
  theme_minimal() +
  labs(color = "Tert_VC")

# Adding the marginal density plots manually
final_plot = ggMarginal(pca_plot, 
                        type = "density", 
                        groupFill = TRUE, 
                        alpha = 0.3)

# Save the final object
ggsave("../../../Results/Figures/Aim_3/Vitamin_C/VC-pathway-pca.png", 
       plot = final_plot, 
       width = 8, 
       height = 6, 
       dpi = 300)
```

### Volcano plot
```{r}
# Join the p-values from Maaslin2 with the log2FC you calculated
plot_data <- daa_annotated_results_df %>%
  left_join(ratio, by = c("feature" = "pathway")) %>%
  mutate(
    # Define significance categories for coloring
    status = case_when(
      p_adjust < 0.05 & log2fc >= 1 ~ "More Abundant (VC Low)",
      p_adjust < 0.05 & log2fc <= -1 ~ "Less Abundant (VC High)",
      TRUE ~ "Not Significant"
    )
  )

# generating volcano plot
volcano_plot <- ggplot(plot_data, aes(x = log2fc, y = -log10(p_adjust), color = status)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("More Abundant (VC Low)" = "#D55E00", 
                                "Less Abundant (VC High)" = "#0072B2", 
                                "Not Significant" = "grey70")) +
  
  theme_minimal() +
  labs(
    x = expression(log[2]~"Fold Change"),
    y = expression(-log[10]~"Adjusted P-value"),
    color = "Significance"
  ) +
  theme(
    text = element_text(size = 20),
    axis.title = element_text(size = 22, face = "bold"),
    axis.text = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.title = element_text(size = 20, face = "bold"),
    legend.position = "bottom"
  )

# Save
ggsave("../../../Results/Figures/Aim_3/Vitamin_C/VC-volcano-plot.png", 
       plot = volcano_plot, width = 12, height = 10, dpi = 300)
```


