---
title: "Vitamin A Picrust differential abundance"
author: "Emily"
date: "2026-03-17"
output: html_document
---

# Introductions
For aim 3 we wanted to identify the functional pathways of significance. This script is designed to identify and visualize functional differences in the gut microbiome between individuals with Low vs. High Vitamin A(VA) intake within a Parkinson’s Disease cohort. It utilizes PICRUst2 pathway predictions and applies Maaslin2 for robust differential abundance testing. Pathway plot, Heatmap plot, and PCA plot is also included here. Note that because the results indicated differences in VA groups were insignificant therefore manual log2fold change checks and volcano plots were not produced. Figures produced here are not included in final manuscript.

## Environment Setup and Data Loading
This section starts the analysis by loading required bioinformatic and visualization libraries and preparing the metadata. 
```{r libraries, include=FALSE}
# Core data manipulation and plotting
library(tidyverse)
library(stringr)

# analysis specific tools
library(ggpicrust2)
library(Maaslin2)

# visualization-specific tools
library(pheatmap)

# Loading metadata and filter for specific VA comparison groups
metadata = read.csv("../../../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

# focusing only on low vs high VA tertiles and drop unused factor levels
metadata_sub = metadata %>% filter(Tert_VE %in% c("VE Low", "VE High")) %>%
  mutate(Tert_VE = fct_drop(as.factor(Tert_VE)))
```

## Differential Abundance Analysis (Maaslin2)
We perform the core statistical testing here, using Maaslin2 to find pathways that differ significantly while adjusting for potential confounders like Sex, Age, and entacapone use.
```{r}

pathway_abundance = read.delim('../../../QIIME2_outputs/path_abun_unstrat.tsv',row.names = 1)

# Filter
path_filt = pathway_abundance %>% select(all_of(rownames(metadata_sub)))


diff_abun = pathway_daa(abundance = path_filt,
                        metadata = metadata,
                        group = "Tert_VE",
                        daa_method = "Maaslin2",
                        fixed_effects = c("Sex", "Age", "entacapone",
                                          "Energy_kcal"),
                        fixed_effects = c("Sex", "Age", "entacapone"),
                        select = NULL,
                        reference = "VE Low")

daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = diff_abun,
                                              ko_to_kegg = FALSE)

# save
saveRDS(daa_annotated_results_df,'daa_annotated_results_df.rds')

readRDS('daa_annotated_results_df.rds')

```
Plots
```{r}
source('../ggpicrust2_errorbar_function_fixed.R')

# Filter out NA in p-values and log2 fold change columns
daa_annotated_results_df_filt = daa_annotated_results_df %>%
  filter(!is.na(p_adjust), !is.na(log2_fold_change))

# Log2 Fold Change values = 8 hits (sufficient for plot)
# Adjusted p-value = 8 hits (non significant)
daa_annotated_results_df_filt %>% 
  filter(p_adjust < 0.09) %>% 
  nrow()

# Log2 Fold Change Values = 19 hits (sufficient for plot)
daa_annotated_results_df_filt %>% 
  filter(abs(log2_fold_change) >= 4e-3) %>% 
  nrow()

# Log2 Fold change values + adjusted p value = 6 hits for plot
daa_annotated_results_df_filt %>% filter(p_adjust< 0.09, abs(log2_fold_change) > 4e-3) %>% nrow()


# Dataframe for plotting
daa_plot <- daa_annotated_results_df %>%
  filter(!is.na(p_adjust), !is.na(log2_fold_change)) %>%
  filter(abs(log2_fold_change) >= 4e-3)


# plotting
peb <- pathway_errorbar_fixed(abundance = path_filt,
                              daa_results_df = daa_plot,
                              Group = metadata_sub$Tert_VE,
                              wrap_label = TRUE,wraplength = 60,
                              fc_cutoff = 4e-3,order_by_log = TRUE,
                              p_values_threshold = 0.09,
                              order = "name",
                              ko_to_kegg = FALSE,
                              p_value_bar = TRUE,
                              x_lab = "description")

# save plot
ggsave("../../../Results/Figures/Aim_3/Vitamin_E/VE-picrust-pathways.png",
       plot = peb, width = 18, height = 10)
```

Heatmap

```{r identical features}
# Heat map library
library('pheatmap')

# working with same set of pathways above
sig_features <- daa_annotated_results_df %>% 
  filter(!is.na(p_adjust), !is.na(log2_fold_change)) %>%
  filter(abs(log2_fold_change) >= 4e-3) %>% 
  pull(feature) %>% 
  unique()

path_relab = path_filt %>% 
  apply(2,function(x) x/sum(x)) %>% as.data.frame()
  colSums(path_relab) # Should be all 1

colSums(path_relab) # Should be all 1


stats = path_relab %>% t() %>%  # switch rows and columns
  as.data.frame() %>% # Lets us use select()
  select(all_of(sig_features)) %>% 
  cor(method = 'pearson') #Pearson tests between every feature pair

# Define the color palette
color_palette <- colorRampPalette(c("blue", "white", "red"))(40)

# Define breaks for the color scale
breaks <- seq(-1, 1, length.out = 41)  # 40 colors + 1 for the endpoint

# Lookup Names
name_lookup <- daa_annotated_results_df %>%
  filter(feature %in% sig_features) %>%
  select(feature, description) %>% # Adjust 'description' to your actual column name
  distinct()

readable_names <- name_lookup$description[match(rownames(stats), name_lookup$feature)]
rownames(stats) <- readable_names
colnames(stats) <- readable_names

view(readable_names)

pheatmap(stats, 
         clustering_distance_rows = "euclidean",  
         clustering_distance_cols = "euclidean", 
         clustering_method = "complete",          
         color = color_palette, 
         breaks = breaks,
         border_color = "white",
         display_numbers = TRUE,
         number_format = "%.2f",      
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 315,              # rotated to minimize white space
         treeheight_row = 25,         
         treeheight_col = 25, 
         legend = TRUE,               
         main = NA,                   
         filename = "../../../Results/Figures/Aim_3/Vitamin_E/MetaCyc-VE_Heatmap_Final-new.png",
         width = 8,                   
         height = 6)

```

PCA plot

```{r PCA}
# Generate pathway PCA plot
pca = pathway_pca(abundance = path_filt,
            metadata = metadata_sub, 
            group = "Tert_VE")

# save
ggsave("../../../Results/Figures/Aim_3/Vitamin_E/VE-pathway-pca.png", 
       plot = pca, 
       width = 8, 
       height = 6, 
       dpi = 300)
```

