---
title: "Manuscript additional plots"
author: Yanting Xiong
output: html_document
date: "2026-04-10"
---

# Introduction
This script produces supplementary visualizations for the final manuscript, specifically focusing on the distribution of clinical cognitive scores (MoCA) within the Parkinson’s Disease cohort to characterize the study population's cognitive health.

# Library loading
Initialize the environment with essential tools for data manipulation and visualization.
```{r library_loading}
library(tidyverse)
library(ggplot2)

```

# Dataset loading
Import processed clinical metadata and calculate the prevalence of cognitive impairment.
```{r dataset_loading}
# Load in the filtered and wrangled Parkinson's metadata
parkinsons_metadata_filt_cat = read.csv("../Datasets/Metadata/parkinsons_metadata_filt_cat.csv", row.names = 1)

parkinsons_metadata_filt_cat_moca_low = parkinsons_metadata_filt_cat %>% 
  filter(moca_total < 26) %>% 
  nrow() # 51 PD patients

```

# Visualize the distribution of MoCA Score using Histogram
Generate a histogram to visualize the spread of total MoCA scores across the cohort.
```{r MoCA_score_visualization}
moca_histogram = ggplot(parkinsons_metadata_filt_cat, aes(x = moca_total)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "black") +
  labs(x = "MoCA Score", y = "Number of PD Patients")
moca_histogram

```